---
title: "First evaluation — multi-sample single-cell nano-CUT&Tag (epigenomic)"
author: "Mattia Zaghi"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
params:
  wd: "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/"
  experiment_source: "discover"   # "discover" (scan wd) or "config" (read config_path)
  config_path: "/home/mattia/Multi_nanoCTRNA/config/config_human_MGI.yaml"
  genome: "hg38"
  binsize: 5000
  filter_passed: true
  qc_low: 0.01
  qc_high: 0.99
  dims: 30
  resolution: 0.6
  workers: 4                      # parallel workers (cores) for independent per-experiment work
  group_pattern: "^(MS|PBMC)"     # samples are joined in one UMAP only within the same matched group
  dataset_label: "this_run"       # label for the current dataset in QC-benchmark plots
  reference_yaml: ""              # optional YAML of external .rds datasets to benchmark against (qc/QC.R style)
  compute_tss: false              # compute TSS enrichment per experiment (heavy) + add to QC
  save_rds: true                  # save the built Seurat objects (.rds) to the result folder
  reuse_counts: true              # reuse cached 5kb count matrices instead of rebuilding them
  fig_scale: 1.25                 # scale factor applied to every saved PNG (bigger so nothing is cut off)
  include_rna: false              # analyse RNA experiments (mark == "RNA") if present
  rna_dims: 15
  rna_res: 0.8
  rna_max_count: 25000
  rna_min_count: 200
  rna_max_mt: 5
  fig_dir: "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval_figures"
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  dev = "png", dpi = 150, fig.align = "center"
)
```

## Overview

This is a **streamlined, sample/modality-agnostic first-evaluation** report. It
covers the first five steps of the standard workflow — load → QC → merge →
normalise/DR → cluster/UMAP — for a quick quality read across all samples at once.
Annotation and label transfer are intentionally out of scope here.

**Nothing is hardcoded to a particular dataset.** By default (`experiment_source =
"discover"`) it scans the analysis folder `params$wd` and picks up *whatever*
`{sample}/{modality}_{barcode}/` experiments are present — add or remove samples on
disk and the report follows. Set `experiment_source = "config"` to instead read the
experiment list from `params$config_path`. RNA experiments (base mark `RNA`) are
analysed only when `include_rna = TRUE`; otherwise the report is epigenomic-only.

Each figure is shown inline **and** written as a high-resolution PNG to
`params$fig_dir`.

```{r setup, results='hide'}
# ---- libraries -------------------------------------------------------------
suppressPackageStartupMessages({
  library(Signac)
  library(Seurat)
  library(GenomicRanges)
  library(stringr)
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
  library(yaml)
  library(ggVennDiagram)
  library(ComplexUpset)
  library(EnsDb.Hsapiens.v86)
  library(BSgenome.Hsapiens.UCSC.hg38)
})

# ---- custom function libraries --------------------------------------------
repodir <- "/home/mattia/Multi_nanoCTRNA/"
source(file.path(repodir, "scripts", "functions_scCT.R"))      # plotCounts, DepthCorMulMod, plotConnectModal, ...
source(file.path(repodir, "scripts", "functions_firstEval.R")) # config-driven first-eval helpers

# ---- parallelism -----------------------------------------------------------
# Independent per-experiment / per-group work is fanned across `params$workers`
# cores. A fork-based (multicore) plan shares the loaded data copy-on-write;
# Signac's own FeatureMatrix/GenomeBinMatrix futures nest as sequential inside.
suppressPackageStartupMessages({ library(future); library(future.apply) })
options(future.globals.maxSize = 128 * 1024^3)   # allow large Seurat objects across workers
if (as.integer(params$workers) > 1 && .Platform$OS.type == "unix") {
  future::plan(future::multicore, workers = as.integer(params$workers))
} else {
  future::plan(future::sequential)
}
cat("Parallel plan:", class(future::plan())[1], "| workers:", future::nbrOfWorkers(), "\n")

# ---- genomic resources -----------------------------------------------------
genome      <- params$genome
genome_ann  <- EnsDb.Hsapiens.v86
bs_genome   <- BSgenome.Hsapiens.UCSC.hg38

# ---- high-resolution figure saver -----------------------------------------
# every saved PNG is scaled up by params$fig_scale so nothing gets cut off.
dir.create(params$fig_dir, recursive = TRUE, showWarnings = FALSE)
fig_scale <- as.numeric(params$fig_scale)
SP <- function(p, nm, w = 8, h = 6)
  save_png(p, nm, params$fig_dir, w * fig_scale, h * fig_scale)
```

## 1. Experiment table (auto-detected)

Experiments are discovered from the analysis folder (or read from the config when
`experiment_source = "config"`) and expanded into one row per
`sample × modality/barcode`. The **base mark** collapses replicate suffixes
(`H3K27ac-1`, `H3K27ac_2` → `H3K27ac`) so replicates can be pooled for the merged
view while remaining distinct experiments. RNA experiments (base mark `RNA`) are
split off and only analysed when `params$include_rna` is `TRUE`.

```{r experiment-table}
exp_all <- build_experiment_table(
  wd          = params$wd,
  config_path = params$config_path,
  source      = params$experiment_source
)

if (is.null(exp_all) || nrow(exp_all) == 0) {
  stop("No experiments found (experiment_source='", params$experiment_source,
       "', wd='", params$wd, "').\n",
       "  wd must be the ANALYSIS ROOT that contains {sample}/{modality}_{barcode}/ dirs",
       " -- not the first_eval/ output subfolder. Run snakemake from that root.")
}

sp      <- split_experiments(exp_all)
exp_df  <- sp$epi            # epigenomic experiments -> bins path (used throughout)
exp_rna <- sp$rna            # RNA experiments -> optional RNA section
samples <- unique(exp_df$sample)
marks   <- unique(exp_df$mark)

knitr::kable(exp_all, caption = "All experiments detected (epigenomic + RNA).")
cat("\nSource:", params$experiment_source,
    "\nSamples:", paste(unique(exp_all$sample), collapse = ", "),
    "\nEpigenomic marks:", paste(marks, collapse = ", "),
    "\nRNA experiments:", nrow(exp_rna),
    if (nrow(exp_rna) > 0 && !params$include_rna)
      "  (present but skipped; set include_rna = TRUE to analyse)" else "",
    "\n")
```

## 2. Load per-cell metadata

Cell-calling metadata (`cell_picking/metadata.csv`) is loaded for every
experiment. When `params$filter_passed` is `TRUE` (default) only cells flagged
`passedMB` by the GMM cell caller are kept. Experiments whose metadata is not yet
on disk are skipped with a warning, so the report can be run while preprocessing
is still finishing.

```{r load-metadata}
meta.ls <- load_metadata_list(exp_df, params$wd, filter_passed = params$filter_passed)
cat("Loaded metadata for", length(meta.ls), "of", nrow(exp_df), "experiments.\n")

# UNFILTERED metadata (all barcodes, passed + dropped) for the cell-selection scatter
meta_all.ls <- load_metadata_list(exp_df, params$wd, filter_passed = FALSE)
cell_all_df <- qc_cell_table(meta_all.ls, group_pattern = params$group_pattern)
groups      <- if (!is.null(cell_all_df)) unique(cell_all_df$group) else character(0)
```

The per-cell QC metric violins and the Excel summary are produced **after** the
bin objects are built, so the fragments-per-cell metric is `nCount_bins` (see the
"Per-cell QC metrics" section below).

# Section A — Cross-sample QC comparison

We compare the per-cell quality metrics across every experiment (sample × mark).
All experiments were loaded into `meta.ls` (one metadata data-frame per `exp_id`),
already restricted to passed cells when `params$filter_passed` is `TRUE`.
Each figure is printed inline and written to `params$fig_dir` as a high-resolution PNG.

## Passed cells per experiment

```{r qc-ncells-barplot, fig.width=9, fig.height=5}
p_ncells <- ncells_barplot(meta.ls)
print(p_ncells)
SP(p_ncells, "QC_ncells_per_experiment", w = 9, h = 5)
```

## Cell selection (passedMB)

The GMM cell-caller decision per experiment: every barcode plotted as unique
fragments (log10) vs FRiP, coloured by whether it was kept (`passedMB TRUE`) or
dropped (`passedMB FALSE`) — the same scatter as the `pick_cells` output. Shown
across all experiments and per sample-type group.

```{r qc-cell-selection, fig.width=12, fig.height=9, results='asis'}
p_sel <- tryCatch(plot_cell_selection(cell_all_df),
                  error = function(e) { message("cell selection: ", conditionMessage(e)); NULL })
if (!is.null(p_sel)) { print(p_sel); SP(p_sel, "QC_cell_selection", w = 12, h = 9) }

for (g in groups) {
  p <- tryCatch(plot_cell_selection(cell_all_df, group = g), error = function(e) NULL)
  if (!is.null(p)) {
    cat("\n\n### Cell selection —", g, "\n\n")
    print(p); SP(p, paste0("QC_cell_selection_", g), w = 12, h = 9)
  }
}
```

## Cell overlap across marks within each sample

For every sample we show how many cells are shared across that sample's marks.
Samples with only one or two marks are handled gracefully (the call is wrapped
in `tryCatch`, so a single degenerate sample never aborts the report).

```{r qc-overlap-per-sample, fig.width=8, fig.height=6, results='asis'}
for (s in samples) {
  cat("\n\n### Overlap — sample:", s, "\n\n")
  tryCatch({
    p_ov <- overlap_plot(meta.ls, s)
    print(p_ov)
    SP(p_ov, paste0("QC_overlap_", s), w = 8, h = 6)
  }, error = function(e) {
    message("Overlap plot skipped for sample '", s, "': ", conditionMessage(e))
  })
}
```

# Build bin × cell objects

For each experiment we build a 5 kb genome-bin × cell matrix from its fragments
file and wrap it in a Seurat/Signac object (assay `bins`). This is the heaviest
step; failures on individual experiments are caught so one bad experiment cannot
abort the report.

```{r build-objects, results='hide'}
# reuse cached 5kb bin matrices (the expensive step) when available, so re-runs
# only build experiments that are new / whose cells changed.
counts_file  <- file.path(dirname(params$fig_dir), "objects",
                          paste0("counts.ls_bins_", params$binsize, ".rds"))
counts_cache <- if (isTRUE(params$reuse_counts) && file.exists(counts_file))
                  readRDS(counts_file) else NULL

obj.ls <- build_bins_objects(
  exp_df, meta.ls, params$wd,
  bs_genome    = bs_genome,
  binsize      = params$binsize,
  min_cell     = 1,
  min_feat     = 1,
  counts_cache = counts_cache
)
```

```{r tss-enrich, eval=params$compute_tss, results='hide'}
# TSS enrichment per experiment (heavy) -- augments each object with a
# TSS.enrichment per-cell column so it flows into the QC metrics below and the
# TSS meta-profile at the end.
ids_o <- names(obj.ls)
obj.ls <- .par_lapply(ids_o, function(id)
  add_tss_enrichment(obj.ls[[id]], genome_ann, genome_name = params$genome))
names(obj.ls) <- ids_o
```

```{r build-objects-summary}
cat("Built", length(obj.ls), "Seurat objects.\n")
data.frame(
  experiment = names(obj.ls),
  n_cells    = vapply(obj.ls, ncol, integer(1)),
  n_bins     = vapply(obj.ls, nrow, integer(1)),
  row.names  = NULL
) |> knitr::kable(caption = "Bin × cell object dimensions per experiment.")
```

```{r save-objects, eval=params$save_rds}
# persist the per-experiment objects + the QC table to the result folder so the
# analysis can be resumed / reused without rebuilding the bin matrices.
rds_dir <- file.path(dirname(params$fig_dir), "objects")   # first_eval/objects/
dir.create(rds_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(obj.ls,  file.path(rds_dir, "obj.ls_bins.rds"))

# raw per-experiment 5kb bin count matrices (sparse) -- the expensive
# GenomeBinMatrix output, saved separately so it can be reused to redo anything.
counts_file <- file.path(rds_dir, paste0("counts.ls_bins_", params$binsize, ".rds"))
saveRDS(extract_counts_list(obj.ls, assay = "bins"), counts_file)
cat("Saved per-experiment objects to:", file.path(rds_dir, "obj.ls_bins.rds"),
    "\n  + count matrices to:", counts_file, "\n")
```

# Per-cell QC metrics (from objects; fragments = `nCount_bins`)

Built from the objects' metadata, so the fragments-per-cell metric is
**`nCount_bins`** (bins-assay counts, as in `qc/QC.R`) — not the metadata
`all_unique_MB`. Metrics: `logUMI`, `nCount_bins`, reads/cell, duplication rate,
FRiP. Shown across all samples and per sample-type group, and exported to
`QC_summary.xlsx`.

```{r qc-cell-df}
cell_df <- qc_cell_table(lapply(obj.ls, function(o) o@meta.data),
                         group_pattern = params$group_pattern)
if (isTRUE(params$save_rds) && !is.null(cell_df)) {
  rds_dir <- file.path(dirname(params$fig_dir), "objects")
  dir.create(rds_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(cell_df, file.path(rds_dir, "qc_cell_df.rds"))
}
```

```{r qc-excel-export}
if (is.null(cell_df) || nrow(cell_df) == 0) {
  cat("No per-cell QC available -- skipping Excel.\n")
} else {
  qc_sum   <- qc_summary_table(cell_df)
  xlsx_out <- file.path(dirname(params$fig_dir), "QC_summary.xlsx")   # first_eval/QC_summary.xlsx
  write_qc_excel(qc_sum, xlsx_out)
  cat("QC summary written to:", xlsx_out, "\n")
  print(knitr::kable(qc_sum, caption = "Per-experiment QC averages (also in QC_summary.xlsx)."))
}
```

```{r qc-metrics-combined, fig.width=10, fig.height=6, results='asis'}
qc_specs <- list(
  list(m = "logUMI",         lab = "log10 UMI",                        log = FALSE),
  list(m = "nCount_bins",    lab = "unique fragments (5kb bins)",      log = TRUE),
  list(m = "passed_filters", lab = "unique fragments (passed_filters)", log = TRUE),
  list(m = "reads",          lab = "reads per cell",                   log = TRUE),
  list(m = "dup_rate",       lab = "duplication rate",                 log = FALSE),
  list(m = "FRiP",           lab = "FRiP (peak_ratio_MB)",             log = FALSE),
  list(m = "TSS_enrichment", lab = "TSS enrichment",                   log = FALSE)
)
for (s in qc_specs) {
  p <- tryCatch(qc_metric_violin(cell_df, s$m, ylab = s$lab, log_y = s$log),
                error = function(e) { message(s$m, ": ", conditionMessage(e)); NULL })
  if (!is.null(p)) { cat("\n\n##", s$lab, "\n\n"); print(p); SP(p, paste0("QC_", s$m), w = 10, h = 6) }
}
```

```{r qc-metrics-by-group, fig.width=9, fig.height=5, results='asis'}
for (g in groups) {
  cat("\n\n## Per-group:", g, "\n\n")
  for (s in qc_specs) {
    p <- tryCatch(qc_metric_violin(cell_df, s$m, ylab = s$lab, log_y = s$log, group = g),
                  error = function(e) NULL)
    if (!is.null(p)) { print(p); SP(p, paste0("QC_", g, "_", s$m), w = 9, h = 5) }
  }
}
```

# Section A2 — QC benchmark (per-cell metrics, `qc/QC.R` style)

The same per-cell QC panels as `qc/QC.R` — **reads/cell**, **fragments/cell**,
**fragment/feature**, **FRiP**, and **duplication rate** — computed on the built
bins objects (which carry both the metadata fields and `nCount_bins`/
`nFeature_bins`). When `params$reference_yaml` points at a references file,
external published datasets are loaded and shown alongside for **cross-technology
comparison**, exactly like `QC.R` (see `config/qc_references_example.yaml`).

```{r qc-benchmark-build}
qc_bench <- qc_benchmark_table(
  obj.ls, exp_df,
  dataset_label  = params$dataset_label,
  reference_yaml = if (nzchar(params$reference_yaml)) params$reference_yaml else NULL,
  group_pattern  = params$group_pattern
)
cat("QC benchmark rows (cells):", nrow(qc_bench),
    "\nDatasets compared:", paste(unique(qc_bench$dataset), collapse = ", "), "\n")
```

```{r qc-benchmark-panels, fig.width=12, fig.height=6, results='asis'}
bench_specs <- list(
  list(m = "reads",          lab = "Reads per cell",                   log = TRUE),
  list(m = "fragments",      lab = "Fragments per cell (bins)",        log = TRUE),
  list(m = "passed_filters", lab = "Unique fragments (passed_filters)", log = TRUE),
  list(m = "frag_per_feat",  lab = "Fragment / feature",               log = FALSE),
  list(m = "frip",           lab = "FRiP (peak_ratio_MB)",             log = FALSE),
  list(m = "dup_rate",       lab = "Duplication rate per cell",        log = FALSE)
)
for (s in bench_specs) {
  p <- tryCatch(qc_benchmark_violin(qc_bench, s$m, ylab = s$lab, log_y = s$log),
                error = function(e) { message(s$m, ": ", conditionMessage(e)); NULL })
  if (!is.null(p)) {
    cat("\n\n###", s$lab, "\n\n"); print(p)
    SP(p, paste0("QCbench_", s$m), w = 12, h = 6)
  }
}
```

### Object-based QC per sample-type group

The same object-based panels (fragments/cell, fragment/feature, …) restricted to
each group, comparing that group's samples on their own axes.

```{r qc-benchmark-by-group, fig.width=9, fig.height=5, results='asis'}
bench_groups <- setdiff(unique(qc_bench$group[qc_bench$dataset == params$dataset_label]), NA)
for (g in bench_groups) {
  cat("\n\n#### Group:", g, "\n\n")
  for (s in bench_specs) {
    p <- tryCatch(qc_benchmark_violin(qc_bench, s$m, ylab = s$lab, log_y = s$log, group = g),
                  error = function(e) NULL)
    if (!is.null(p)) { print(p); SP(p, paste0("QCbench_", g, "_", s$m), w = 9, h = 5) }
  }
}
```

## TSS enrichment (optional)

Runs only when `params$compute_tss = TRUE` (it is heavy: it annotates each object
from the EnsDb and scans fragments around TSSs).

```{r qc-tss, eval=params$compute_tss, fig.width=12, fig.height=4, results='asis'}
# TSS was already computed on obj.ls above; here we just draw the meta-profiles.
for (id in names(obj.ls)) {
  p <- tryCatch(Signac::TSSPlot(obj.ls[[id]]) + ggplot2::ggtitle(id),
                error = function(e) { message("TSSPlot failed for ", id, ": ", conditionMessage(e)); NULL })
  if (!is.null(p)) { cat("\n\n###", id, "\n\n"); print(p); SP(p, paste0("TSS_", id), w = 8, h = 4) }
}
```

# Section B — Normalisation, dimensionality reduction, UMAP & clustering

UMAPs are built two ways: **jointly per _sample type_ × mark** (samples are merged
only within the same group — e.g. all MS H3K27ac together, all PBMC H3K27ac
together — **never mixing types**, controlled by `params$group_pattern`), and
**per individual experiment**. Bins objects are normalised with TF-IDF, top
features selected, and an LSI (`RunSVD`) embedding computed. The first LSI
component is almost always dominated by sequencing depth, so all downstream
neighbour/UMAP steps start at LSI dimension 2 (`dims = 2:params$dims`).
Independent group/experiment work is fanned across `params$workers` cores.

## Merge by sample-type × mark, and normalise

```{r b-merge-normalize}
merged.ls <- merge_by_group_mark(obj.ls, exp_df, group_pattern = params$group_pattern)
cat("Joint groups (type x mark):", paste(names(merged.ls), collapse = ", "), "\n")

grp_names <- names(merged.ls)
merged.ls <- .par_lapply(grp_names, function(nm) {
  tryCatch(
    normalize_dr(merged.ls[[nm]], min_cutoff = 5),
    error = function(e) {
      message("normalize_dr failed for '", nm, "': ", conditionMessage(e))
      merged.ls[[nm]]
    }
  )
})
names(merged.ls) <- grp_names
```

## Depth correlation of LSI components

`DepthCorMulMod()` shows the correlation between each LSI component and
sequencing depth; the strong dim-1 correlation is the reason it is dropped.

```{r b-depthcor, fig.width=12, fig.height=8}
depth_plots <- lapply(names(merged.ls), function(nm) {
  tryCatch(DepthCorMulMod(merged.ls[[nm]]),
           error = function(e) NULL)
})
depth_plots <- Filter(Negate(is.null), depth_plots)
if (length(depth_plots) > 0) {
  p_depth <- ggpubr::ggarrange(plotlist = depth_plots,
                               ncol = min(2, length(depth_plots)),
                               nrow = ceiling(length(depth_plots) / 2))
  print(p_depth)
  SP(p_depth, "DR_depthcor_per_mark", w = 12, h = 8)
}
```

## Elbow plots

Standard-deviation explained per LSI dimension, used to sanity-check the number
of dimensions retained.

```{r b-elbow, fig.width=12, fig.height=8}
elbow_plots <- lapply(names(merged.ls), function(nm) {
  tryCatch({
    obj  <- merged.ls[[nm]]
    ndim <- min(50, ncol(obj[["lsi"]]))
    ElbowPlot(obj, reduction = "lsi", ndims = ndim) +
      ggplot2::ggtitle(nm) +
      ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, face = "bold"))
  }, error = function(e) NULL)
})
elbow_plots <- Filter(Negate(is.null), elbow_plots)
if (length(elbow_plots) > 0) {
  p_elbow <- ggpubr::ggarrange(plotlist = elbow_plots,
                               ncol = min(2, length(elbow_plots)),
                               nrow = ceiling(length(elbow_plots) / 2))
  print(p_elbow)
  SP(p_elbow, "DR_elbow_per_mark", w = 12, h = 8)
}
```

## UMAP and clustering (joint per sample-type × mark)

```{r b-umap-cluster}
cur_names <- names(merged.ls)
merged.ls <- .par_lapply(cur_names, function(nm) {
  tryCatch(
    run_umap_cluster(merged.ls[[nm]],
                     dims = 2:params$dims,
                     res  = params$resolution),
    error = function(e) {
      message("run_umap_cluster failed for '", nm, "': ", conditionMessage(e))
      merged.ls[[nm]]
    }
  )
})
names(merged.ls) <- cur_names
```

```{r save-merged, eval=params$save_rds}
rds_dir <- file.path(dirname(params$fig_dir), "objects")
dir.create(rds_dir, recursive = TRUE, showWarnings = FALSE)

# one .rds per sample-type group (MS, PBMC, PBMCs_nanoCTR, ...), each a named
# list of that group's per-mark merged objects (names = mark).
grp_of <- sub("\\|.*$", "", names(merged.ls))            # "MS|H3K27ac" -> "MS"
saved <- character(0)
for (g in unique(grp_of)) {
  sub_ls <- merged.ls[grp_of == g]
  names(sub_ls) <- sub("^[^|]*\\|", "", names(sub_ls))   # keep just the mark as the name
  safe <- gsub("[^A-Za-z0-9]+", "_", g)
  f <- file.path(rds_dir, paste0("merged_", safe, ".rds"))
  saveRDS(sub_ls, f)
  saved <- c(saved, basename(f))
}
saveRDS(merged.ls, file.path(rds_dir, "merged.ls_group_mark.rds"))  # combined, for convenience
cat("Saved per-group merged objects:", paste(saved, collapse = ", "),
    "\n  (+ combined merged.ls_group_mark.rds)\n")
```

For each sample-type × mark we show the clustering (left) next to the sample of
origin (right); samples of the same type should mix within clusters.

```{r b-dimplots, fig.width=12, fig.height=6, results='asis'}
for (nm in names(merged.ls)) {
  cat("\n\n###", nm, "\n\n")
  tryCatch({
    obj <- merged.ls[[nm]]
    if (!"umap" %in% names(obj@reductions)) {
      message("No UMAP for mark '", nm, "', skipping.")
      next
    }
    p_clu <- DimPlot(obj, reduction = "umap", label = TRUE) +
      NoLegend() + ggplot2::ggtitle(nm)
    p_smp <- DimPlot(obj, reduction = "umap", group.by = "sample") +
      ggplot2::ggtitle("sample")

    nm_safe <- gsub("[^A-Za-z0-9]+", "_", nm)
    print(p_clu)
    print(p_smp)
    SP(p_clu, paste0("UMAP_", nm_safe, "_clusters"), w = 7, h = 6)
    SP(p_smp, paste0("UMAP_", nm_safe, "_bySample"), w = 7, h = 6)

    # colour by patient when it adds information beyond sample (e.g. MS pools)
    if ("patient" %in% colnames(obj@meta.data) &&
        length(unique(obj$patient)) > 1 &&
        !setequal(unique(obj$patient), unique(obj$sample))) {
      p_pat <- DimPlot(obj, reduction = "umap", group.by = "patient") +
        ggplot2::ggtitle("patient")
      print(p_pat)
      SP(p_pat, paste0("UMAP_", nm_safe, "_byPatient"), w = 7, h = 6)
    }

    p_both <- ggpubr::ggarrange(p_clu, p_smp, ncol = 2, nrow = 1)
    print(p_both)
    SP(p_both, paste0("UMAP_", nm_safe, "_clusters_and_sample"), w = 12, h = 6)
  }, error = function(e) {
    message("DimPlot failed for mark '", nm, "': ", conditionMessage(e))
  })
}
```

## Per-sample (single-experiment) UMAPs

The same pipeline (`normalize_dr` → `run_umap_cluster`) is run on each
individual experiment so per-sample structure can be inspected before merging.
Experiments with fewer than 50 cells are skipped, and every per-object step is
guarded so one bad experiment cannot abort the report. Results are arranged into
one grid per mark.

```{r b-per-experiment-umap, fig.width=12, fig.height=8, results='asis'}
# Heavy compute (normalise + UMAP per experiment) fanned across cores; the
# resulting DimPlots are collected and arranged serially below.
per_exp <- .par_lapply(names(obj.ls), function(id) {
  o <- obj.ls[[id]]
  if (is.null(o) || ncol(o) < 50) return(NULL)
  tryCatch({
    o <- normalize_dr(o, min_cutoff = 5)
    o <- run_umap_cluster(o, dims = 2:params$dims, res = params$resolution)
    DimPlot(o, reduction = "umap", label = TRUE) + NoLegend() + ggplot2::ggtitle(id)
  }, error = function(e) {
    message("Per-experiment UMAP failed for '", id, "': ", conditionMessage(e)); NULL
  })
})
names(per_exp) <- names(obj.ls)

per_exp_umaps <- list()
for (id in names(per_exp)) {
  if (is.null(per_exp[[id]])) next
  mk <- exp_df$mark[match(id, exp_df$exp_id)]
  if (is.na(mk)) mk <- "unknown"
  per_exp_umaps[[mk]] <- c(per_exp_umaps[[mk]], setNames(list(per_exp[[id]]), id))
}

for (mk in names(per_exp_umaps)) {
  cat("\n\n### Per-sample UMAPs — mark:", mk, "\n\n")
  plist <- per_exp_umaps[[mk]]
  p_grid <- ggpubr::ggarrange(plotlist = plist,
                              ncol = min(2, length(plist)),
                              nrow = ceiling(length(plist) / 2))
  print(p_grid)
  SP(p_grid, paste0("UMAP_perSample_", mk), w = 12, h = 8)
}
```

# Section C — RNA (optional)

Run only when `params$include_rna` is `TRUE` **and** RNA experiments (base mark
`RNA`) were detected. RNA is loaded from the cellranger GEX
`filtered_feature_bc_matrix.h5`, QC-filtered on counts / mitochondrial content,
then log-normalised, PCA-reduced, merged across samples, clustered and embedded
with UMAP. When RNA is absent or `include_rna = FALSE`, these chunks do not run.

```{r c-rna-load, eval=params$include_rna}
have_rna <- nrow(exp_rna) > 0
if (!have_rna)
  message("include_rna = TRUE but no RNA experiments detected; skipping RNA section.")
rna.obj.ls <- if (have_rna) load_rna_objects(exp_rna, params$wd) else list()
cat("Loaded", length(rna.obj.ls), "RNA object(s).\n")
```

```{r c-rna-qc, eval=params$include_rna, fig.width=9, fig.height=5}
if (length(rna.obj.ls) > 0) {
  rna_meta.ls <- lapply(rna.obj.ls, function(o) o@meta.data)
  for (feat in c("nCount_RNA", "nFeature_RNA", "percent.MT")) {
    p <- tryCatch(qc_violin(rna_meta.ls, feat, ylab = feat,
                            log_y = feat == "nCount_RNA"),
                  error = function(e) NULL)
    if (!is.null(p)) { print(p); SP(p, paste0("RNA_QC_", feat), w = 9, h = 5) }
  }
}
```

```{r c-rna-cluster, eval=params$include_rna, fig.width=12, fig.height=6, results='asis'}
if (length(rna.obj.ls) > 0) {
  # QC filter each experiment, then merge across samples and cluster/UMAP
  rna.obj.ls <- lapply(rna.obj.ls, function(o)
    tryCatch(qc_filter_rna(o, params$rna_max_count, params$rna_min_count,
                           params$rna_max_mt),
             error = function(e) o))
  rna_merged <- tryCatch(merge_by_mark(rna.obj.ls, exp_rna),
                         error = function(e) list())
  for (nm in names(rna_merged)) {
    obj <- tryCatch({
      o <- normalize_rna(rna_merged[[nm]])
      run_umap_cluster_rna(o, dims = 1:params$rna_dims, res = params$rna_res)
    }, error = function(e) {
      message("RNA DR/clustering failed for '", nm, "': ", conditionMessage(e)); NULL
    })
    if (is.null(obj)) next
    cat("\n\n### RNA:", nm, "\n\n")
    p_clu <- DimPlot(obj, reduction = "umap", label = TRUE) +
      NoLegend() + ggplot2::ggtitle(nm)
    p_smp <- DimPlot(obj, reduction = "umap", group.by = "sample") +
      ggplot2::ggtitle("sample")
    print(p_clu); print(p_smp)
    SP(p_clu, paste0("RNA_UMAP_", nm, "_clusters"), w = 7, h = 6)
    SP(p_smp, paste0("RNA_UMAP_", nm, "_bySample"), w = 7, h = 6)
  }
}
```

# Session information

```{r session-info}
cat("Figures written to:", params$fig_dir, "\n\n")
sessionInfo()
```
