---
title: "First evaluation — interactive"
author: "Mattia Zaghi"
output:
  html_document:
    toc: true
    toc_float: true
---

Interactive companion to `first_eval_MGI.Rmd`. Unlike the report (which is
parametrised and meant to be rendered by snakemake), this one **defines its
config in a plain chunk** so you can run it **chunk-by-chunk in RStudio**
(Ctrl/Cmd+Enter) and continue the analysis. It **loads the objects the report
already saved** — no rebuilding.

Needs an R with the analysis packages (Signac/Seurat/...): your usual RStudio R,
or `conda activate nanoctarna-analysis; rstudio`.

```{r setup, message=FALSE, warning=FALSE}
# ---- libraries + helper functions + genome --------------------------------
suppressPackageStartupMessages({
  library(Signac); library(Seurat); library(GenomicRanges)
  library(stringr); library(dplyr); library(ggplot2); library(ggpubr)
  library(yaml); library(ggVennDiagram); library(ComplexUpset)
  library(EnsDb.Hsapiens.v86); library(BSgenome.Hsapiens.UCSC.hg38)
})
repodir <- "/home/mattia/Multi_nanoCTRNA"
source(file.path(repodir, "scripts", "functions_scCT.R"))       # plotCounts, DepthCorMulMod, ...
source(file.path(repodir, "scripts", "functions_firstEval.R"))  # config-driven first-eval helpers

genome     <- "hg38"
genome_ann <- EnsDb.Hsapiens.v86
bs_genome  <- BSgenome.Hsapiens.UCSC.hg38
```

## Config (edit if your run is elsewhere)

`params` is a plain list here (not YAML), so every chunk that uses `params$...`
— including chunks copied from `first_eval_MGI.Rmd` — works when run interactively.

```{r config}
wd         <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/"   # analysis root
result_dir <- file.path(wd, "first_eval")
obj_dir    <- file.path(result_dir, "objects")

params <- list(
  wd            = paste0(wd, "/"),
  binsize       = 5000,
  filter_passed = TRUE,
  dims          = 30,
  resolution    = 0.6,
  workers       = 4,
  group_pattern = "^(PBMCs)",#"^(MS|PBMC)", now only PBMCs
  fig_dir       = file.path(result_dir, "figures"),
  fig_scale     = 1.25
)
dir.create(params$fig_dir, recursive = TRUE, showWarnings = FALSE)
fig_scale <- params$fig_scale
SP <- function(p, nm, w = 8, h = 6)                     # same figure saver as the report
  save_png(p, nm, params$fig_dir, w * fig_scale, h * fig_scale)
```

## Load what the report already built

```{r load-objects}
load_if <- function(f) if (file.exists(f)) readRDS(f) else {
  message("not found (run the report first): ", f); NULL }

obj.ls    <- load_if(file.path(obj_dir, "obj.ls_bins.rds"))            # list keyed by experiment
#merged.ls <- load_if(file.path(obj_dir, "merged.ls_group_mark.rds"))  # list keyed by "group|mark"
#cell_df   <- load_if(file.path(obj_dir, "qc_cell_df.rds"))            # per-cell QC table
# also available individually (per group), and the raw 5kb matrices:
#   merged_MS   <- load_if(file.path(obj_dir, "merged_MS.rds"))       # list(H3K27ac=, H3K27me3=)
#   merged_PBMC <- load_if(file.path(obj_dir, "merged_PBMC.rds"))
#   counts.ls   <- load_if(file.path(obj_dir, paste0("counts.ls_bins_", params$binsize, ".rds")))

cat("obj.ls   :", if (is.null(obj.ls))    "-" else paste(length(obj.ls), "experiments"), "\n",
    "merged.ls:", if (is.null(merged.ls)) "-" else paste(names(merged.ls), collapse = ", "), "\n",
    "cell_df  :", if (is.null(cell_df))   "-" else paste(nrow(cell_df), "cells"), "\n")
```

## Doublet removal (AMULET) -- must run before merge/cluster, on obj.ls

`obj.ls` still has one Signac object per experiment with native (unprefixed)
barcodes -- exactly what AMULET needs, since it scores multiplets from a
single experiment's `fragments_raw.tsv.gz`. We flag+drop doublets here, then
rebuild `merged.ls` from the cleaned `obj.ls` (cheap: `merge_by_group_mark`
just re-merges already-computed objects, no matrix rebuild).

```{r doublets_amulet}
amulet_dir <- "/home/mattia/AMULET"                                   # AMULET.py location
chrlist    <-  file.path(amulet_dir, "ref/chrName.txt")        # TODO: add to config/ (hg38)
blacklist  <- file.path(amulet_dir, "ref/hg38.blacklist.bed") # TODO: add to config/ (hg38)

# AMULET needs pandas/numpy/pysam -- neither nanoscope_general nor
# nanoscope_debarcode ship pandas, so it gets its own tiny env
# (envs/nanoscope_amulet.yaml): `conda env create -f envs/nanoscope_amulet.yaml`.
# system() spawns a plain subshell that does NOT re-source your `conda activate`,
# so point it at the env's python by absolute path rather than relying on
# whatever "python" happens to be on PATH when this is run from RStudio.
amulet_python <- "/home/mattia/miniconda3_n//envs/nanoscope_amulet/bin/python3"  # adjust to your conda prefix
stopifnot("amulet_python not found -- fix the path or `conda env create -f envs/nanoscope_amulet.yaml`" =
            file.exists(amulet_python))

# exp_id -> fragments file, without assuming barcode format. Splits on the
# FIRST "." (sample never contains one; modality might in principle, but
# doesn't in this config) and globs <wd>/<sample>/<modality>_*/... for the
# cellranger outs dir, so it matches "H3K27ac_MS058", "H3K27ac_CCTATCCT", etc.
find_fragments <- function(exp_id, wd) {
  sample   <- sub("\\..*$", "", exp_id)
  modality <- sub("^[^.]+\\.", "", exp_id)
  exp_dir  <- Sys.glob(file.path(wd, sample, paste0(modality, "_*")))
  if (length(exp_dir) == 0) return(NA_character_)
  cand <- Sys.glob(file.path(exp_dir, "cellranger", "outs",
                             c("fragments_raw.tsv.gz", "fragments.tsv.gz")))
  if (length(cand) == 0) return(NA_character_)
  cand[1]
}

# AMULET is a two-step CLI, not one call (there is no --forcesorted flag on
# either script -- that's what the earlier run errored on):
#   1) FragmentFileOverlapCounter.py <fragments> <bcmap.csv> <chrlist> <outdir>
#      -> writes Overlaps.txt / OverlapSummary.txt into <outdir>. <bcmap.csv>
#      must be a csv *with a header row*, columns "barcode" and
#      "is__cell_barcode" (rows kept where == 1) -- not the one-barcode-
#      per-line file this chunk used to write.
#   2) AMULET.py --rfilter <blacklist.bed> <Overlaps.txt> <OverlapSummary.txt> <outdir2>
#      -> the repeat blacklist is only ever an AMULET.py argument
#      (--rfilter); it does not go into step 1 at all.
flag_doublets_amulet <- function(exp_id, obj, wd, qval_cutoff = 0.05) {
  frag <- find_fragments(exp_id, wd)
  if (is.na(frag)) { message("no fragments for ", exp_id, ", skipping"); return(obj) }

  counter_dir <- file.path(obj_dir, "amulet", exp_id, "overlaps")
  amulet_out  <- file.path(obj_dir, "amulet", exp_id, "multiplets")
  dir.create(counter_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(amulet_out,  recursive = TRUE, showWarnings = FALSE)

  # BCMAP: header "barcode,is__cell_barcode" -- score only cells that passed
  # pick_cells.R, so every row here is a valid cell (is__cell_barcode = 1)
  bc_file <- file.path(counter_dir, "singlecell.csv")
  write.csv(data.frame(barcode = colnames(obj), is__cell_barcode = 1L,
                        check.names = FALSE),
            bc_file, row.names = FALSE, quote = FALSE)

  # step 1: per-fragment overlap counting (no --rfilter here)
  system(sprintf("%s %s/FragmentFileOverlapCounter.py %s %s %s %s",
                  amulet_python, amulet_dir, frag, bc_file, chrlist, counter_dir))

  overlaps_path <- file.path(counter_dir, "Overlaps.txt")
  summary_path  <- file.path(counter_dir, "OverlapSummary.txt")
  if (!file.exists(overlaps_path) || !file.exists(summary_path)) {
    message("FragmentFileOverlapCounter produced no output for ", exp_id); return(obj)
  }

  # step 2: multiplet probability estimation; blacklist goes here via --rfilter
  system(sprintf("%s %s/AMULET.py --rfilter %s %s %s %s",
                  amulet_python, amulet_dir, blacklist, overlaps_path, summary_path, amulet_out))

  probs_path <- file.path(amulet_out, "MultipletProbabilities.txt")
  if (!file.exists(probs_path)) { message("AMULET produced no output for ", exp_id); return(obj) }

  # AMULET.py writes columns cell_id, barcode, p-value, q-value and already
  # BH-corrects q-value itself -- no need to re-run p.adjust. Match by
  # barcode explicitly rather than assuming row order/rownames.
  probs <- read.delim(probs_path, check.names = FALSE)
  qval  <- setNames(probs[["q-value"]], probs[["barcode"]])[colnames(obj)]
  obj$amulet_qval       <- qval
  obj$is_doublet_amulet <- qval < qval_cutoff
  obj
}

obj.ls <- Map(flag_doublets_amulet, names(obj.ls), obj.ls, MoreArgs = list(wd = params$wd))
saveRDS(obj.ls,"/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects/obj.ls_bins_Amulet.rds")
n_scored   <- sum(sapply(obj.ls, function(o) !is.null(o$is_doublet_amulet)))
n_doublets <- sum(sapply(obj.ls, function(o) if (is.null(o$is_doublet_amulet)) 0 else sum(o$is_doublet_amulet, na.rm = TRUE)))
cat(n_doublets, "multiplets flagged across", n_scored, "of", length(obj.ls), "experiments",
    "(the rest had no fragments file found -- check find_fragments() against your layout)\n")

# drop flagged cells from obj.ls (native names) and from merged.ls (which may
# have prefixed names "<exp_id>_<barcode>" only for marks merged from >1 exp)
for (id in names(obj.ls)) {
  o <- obj.ls[[id]]
  if (is.null(o$is_doublet_amulet)) next
  bad <- colnames(o)[o$is_doublet_amulet]
  if (length(bad) == 0) next
  obj.ls[[id]] <- subset(o, cells = colnames(o)[!o$is_doublet_amulet])
  for (nm in names(merged.ls)) {
    m    <- merged.ls[[nm]]
    drop <- intersect(colnames(m), c(bad, paste0(id, "_", bad)))
    if (length(drop) > 0) merged.ls[[nm]] <- subset(m, cells = setdiff(colnames(m), drop))
  }
}
```
create a merged MS object

```{r merge_modalities}

merged.ls <- merge_by_group_mark(obj.ls, exp_df, group_pattern = params$group_pattern)
cat("Joint groups (type x mark):", paste(names(merged.ls), collapse = ", "), "\n")

grp_names <- names(merged.ls)
merged.ls <- .par_lapply(grp_names, function(nm) {
  tryCatch(
    normalize_dr(merged.ls[[nm]], min_cutoff = 5),
    error = function(e) {
      message("normalize_dr failed for '", nm, "': ", conditionMessage(e))
      merged.ls[[nm]]
    }
  )
})
names(merged.ls) <- grp_names
```




```{r batch_harmony}

suppressPackageStartupMessages(library(harmony))

# 1) technology, from sample
merged.ls$`H3K27ac`@meta.data$technology <- case_when(
  merged.ls$`H3K27ac`@meta.data$sample %in% c("MS3BL1_CT", "MS3BL2_CT") ~ "GEM-X",
  merged.ls$`H3K27ac`@meta.data$sample  == "MS549_CT"                ~ "Next-GEM",
  TRUE                                     ~ NA_character_
)

merged.ls$`H3K27me3`$technology <- case_when(
  merged.ls$`H3K27me3`@meta.data$sample %in% c("MS3BL1_CT", "MS3BL2_CT") ~ "GEM-X",
  merged.ls$`H3K27me3`@meta.data$sample  == "MS549_CT"                ~ "Next-GEM",
  TRUE                                     ~ NA_character_
)


# 2) protocol, from sample AND the technology just derived above
merged.ls$`H3K27ac`@meta.data$protocol <- case_when(
  merged.ls$`H3K27ac`@meta.data$patient %in% c("MS058CT1", "MS058CT2", "MS381CT1","MS381CT2","MS549_CT") ~ "fresh",
  merged.ls$`H3K27ac`@meta.data$patient == "MS549CT" ~ "frozen",
  TRUE                                                                        ~ NA_character_
)

merged.ls$`H3K27me3`@meta.data$protocol <- case_when(
  merged.ls$`H3K27me3`@meta.data$patient %in% c("MS058CT1", "MS058CT2", "MS381CT1","MS381CT2","MS549_CT") ~  "fresh" ,
  merged.ls$`H3K27me3`@meta.data$patient == "MS549CT"~"frozen",
  TRUE                                                                        ~ NA_character_
)


merged.ls <- lapply(merged.ls, function(obj) {
  #obj <- RunTFIDF(obj) > FindTopFeatures(min.cutoff = 5) > RunSVD()
  obj <- RunHarmony(obj, group.by.vars = c("technology","protocol"),
                    reduction.use = "lsi", reduction.save = "harmony_lsi",
                    project.dim = FALSE)
  obj <- RunUMAP(obj, reduction = "harmony_lsi", di = 2:params$dims, reduction.name = "umap_harmony")
  obj <- FindNeighbors(obj, reduction = "harmony_lsi", di = 2:params$dims)
  obj <- FindClusters(obj, resolution = params$resolution, algorithm = 3, verbose = FALSE,cluster.name="harmony_clusters",graph.name = "bins_snn")
  obj
})
```


```{r umaps, fig.width=12, fig.height=6}
nm  <- "MS|H3K27ac"                       # pick any name from names(merged.ls)
obj <- merged.ls[[nm]]

p_clu <- DimPlot(obj, reduction = "umap_harmony", label = TRUE) + NoLegend() + ggtitle(nm)
p_smp <- DimPlot(obj, reduction = "umap_harmony", group.by = "sample")  + ggtitle("sample")
p_pat <- DimPlot(obj, reduction = "umap_harmony", group.by = "patient") + ggtitle("patient")
ggarrange(p_clu, p_smp, p_pat, ncol = 3)

```

```{r recluster}
# re-cluster / re-embed with different parameters and inspect
obj <- FindClusters(obj, resolution = 0.4, algorithm = 3, verbose = FALSE)
DimPlot(obj, reduction = "umap", label = TRUE) + NoLegend()
```

## Continue: QC of any metric (from the saved cell_df)

```{r qc}
# metrics available: logUMI, nCount_bins, passed_filters, reads, dup_rate, FRiP, TSS_enrichment
qc_metric_violin(cell_df, "nCount_bins", ylab = "unique fragments (bins)", log_y = TRUE)
# per group:
# qc_metric_violin(cell_df, "TSS_enrichment", ylab = "TSS enrichment", group = "MS")
```

## (Optional) rebuild instead of loading

Only if you changed something upstream. Reuses the cached 5kb matrices, so it's fast.

```{r rebuild, eval=FALSE}
exp_df  <- discover_experiments(params$wd)
meta.ls <- load_metadata_list(exp_df, params$wd, filter_passed = params$filter_passed)
obj.ls  <- build_bins_objects(
  exp_df, meta.ls, params$wd, bs_genome = bs_genome, binsize = params$binsize,
  counts_cache = readRDS(file.path(obj_dir, paste0("counts.ls_bins_", params$binsize, ".rds"))))
```

## Create gene activity for label transfer with Mukund Dataset
### 6.1 Add gene annotations to the Seurat object
Before creating a gene activity matrix we first need to add gene annotations to the H3K27ac Seurat object.
```{r add_gene_ann }
# get gene annotations from Ensembl
annotations <- GetGRangesFromEnsDb(ensdb = genome_ann)

# change to UCSC style since the data was mapped to mm10
seqlevelsStyle(annotations) <- 'UCSC'

# add the gene information to the object
Annotation(merged.ls$`MS|H3K27ac`) <- annotations
```

```{r gene_activity}

# Calculate gene activities
gene.activities <- GeneActivity(merged.ls$`MS|H3K27ac`)

# Add the gene activity matrix to the Seurat object as a new assay
merged.ls$`MS|H3K27ac`[['RNA']] <- CreateAssayObject(counts = gene.activities)
# Set RNA as default assay
DefaultAssay(merged.ls$`MS|H3K27ac`) <- 'RNA'
# normalize RNA data
merged.ls$`MS|H3K27ac` <- NormalizeData(merged.ls$`MS|H3K27ac`) |>
            FindVariableFeatures() |>
            ScaleData()

```

```{r}
Human_H3K27ac_nanoCT <- readRDS("/date/gcb/gcb_wq/MS_project/reference_datasets/Kabbe_2026_nCT_rds/processed_data/H3K27ac_GRS_anno_EA.rds")

DefaultAssay(Human_H3K27ac_nanoCT) <- 'PROM2KBGB'
  
feats   <- intersect(VariableFeatures(MS$H3K27ac@assays$RNA), rownames(Human_H3K27ac_nanoCT))
 
anchors <- FindTransferAnchors(reference = Human_H3K27ac_nanoCT, query = MS$H3K27ac,
               reference.assay = "PROM2KBGB", query.assay = "RNA",
               reduction = "cca", dims = 1:50, features = feats)

MS$H3K27ac <- MapQuery(
  anchorset =anchors, 
  reference = Human_H3K27ac_nanoCT,
  query =   MS$H3K27ac,
  refdata = "transferred_annot")
 
#rename predictions columns 
cn <- colnames(MS$H3K27ac@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_CZI_Human"
cn[cn == "predicted.id.score"] <- "predicted.id.score_CZI_Human"
colnames(MS$H3K27ac@meta.data) <- cn


```


```{r}
Human_MS<- readRDS("/date/gcb/gcb_YD/MS_sc_project/MSref_brain_EA/ms_lesions_snRNAseq_seuratV5_WMEA_updated_fullFineLesionDown500.rds")
new_names <- sub("-ENSG[0-9]+$", "", rownames(Human_MS))
sum(duplicated(new_names))

DefaultAssay(Human_MS) <- 'RNA'


for (a in Assays(Human_MS)) {
  rn <- rownames(Human_MS[[a]])
  new_rn <- sub("-ENSG[0-9]+$", "", rn)
  if (any(duplicated(new_rn))) new_rn <- make.unique(new_rn)
  rownames(Human_MS[[a]]) <- new_rn
}


feats   <- intersect(VariableFeatures(merged.ls$`MS|H3K27ac`@assays$RNA), rownames(Human_MS))
 
anchors <- FindTransferAnchors(reference = Human_MS, query = merged.ls$`MS|H3K27ac`,
               reference.assay = "RNA", query.assay = "RNA",
               reduction = "cca", dims = 1:50, features = feats)

merged.ls$`MS|H3K27ac` <- MapQuery(
  anchorset =anchors, 
  reference = Human_MS,
  query =   merged.ls$`MS|H3K27ac`,
  refdata = "type_broad")
 
#rename predictions columns 
cn <- colnames(merged.ls$`MS|H3K27ac`@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Human_MS"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Human_MS"
colnames(merged.ls$`MS|H3K27ac`@meta.data) <- cn

```

```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/MS_metadata.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@meta.data
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/MS_UMAP_coordinates_harmony.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@reductions[["umap_harmony"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/MS_UMAP_coordinates.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@reductions[["umap"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")
```

Label transfer PBMCs with single cell ATAC


```{r}

suppressPackageStartupMessages({
  library(data.table)
  library(Signac)
  library(Seurat)
  library(GenomicRanges)
  library(EnsDb.Hsapiens.v86)             # swap for the matching EnsDb if not hg38
  library(BSgenome.Hsapiens.UCSC.hg38)    # swap for the matching BSgenome if not hg38 -- only used for chrom sizes to tile
})

set.seed(1)

# ---- 1. user-editable paths ------------------------------------------------
meta_path <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/Refs/GSE306525_cell_annotations.tsv"   # <- point at your actual TSV
genome_build <- "hg38"

# GEO series crosswalk (built once from GSE306525_series_matrix.txt's
# Sample_geo_accession/_title/_supplementary_file_1 rows -- see
# scripts/geo_sample_crosswalk.csv) and the local directory holding the
# downloaded GEO fragment files.
crosswalk_path <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/Refs/geo_sample_crosswalk.csv"
fragments_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/Refs"

# ---- 2. read + reshape metadata -------------------------------------------
meta <- fread(meta_path)

# Split "{Sample}#{barcode}" -> archr_sample / barcode (bare 10x barcode,
# which is what has to match the fragments file's 4th column).
meta[, c("archr_sample", "barcode") := tstrsplit(cellId_archr, "#", fixed = TRUE)]
stopifnot(all(meta$archr_sample == meta$Sample))  # sanity check the split

# Keep only cells ArchR already called as real cells.
meta <- meta[PassQC == 1]

# ArchR records whatever local file name was fed in as the `Sample` value
# verbatim when the Arrow files were created, so the string doesn't
# necessarily match the bare GEO donor code one-for-one -- your metadata's
# samples carry the full original local filename including extension (e.g.
# "ATAC_055_fragments.tsv.gz", confirmed by the earlier missing-file error,
# which showed even "ATAC_EV08_fragments.tsv.gz" in that form). Normalize by
# stripping a trailing "_fragments.<ext>.gz" / ".<ext>.gz" before comparing
# against anything bare (HEALTHY_SAMPLES below, or geo_sample_crosswalk.csv's
# `archr_sample_guess`), but keep working with the *raw*, un-stripped Sample
# string everywhere else, since that's what the rest of the script indexes
# with (sample_map[[s]], `s` coming straight from `unique(meta$archr_sample)`).
strip_suffix <- function(x) sub("(_fragments)?\\.(tsv|bed)\\.gz$", "", x)

# ---- 2a. restrict to only the samples you actually have fragments for ------
# Your metadata spans the full GSE306525 series (C19 severity panel, flu
# time-course, ...), but you only want the object built from the six C19
# healthy-control (C19_ctrl) donors whose fragments.bed.gz you've downloaded.
# Everything else is dropped here so the crosswalk/existence checks below
# never even look at the samples you don't have.
HEALTHY_SAMPLES <- c(
  "ATAC_HIP02_frozen",
  "ATAC_HIP023_frozen",
  "ATAC_HIP043",
  "ATAC_HIP044",
  "ATAC_HIP045",
  "ATAC_HIP15_frozen"
)
meta[, archr_sample_bare := strip_suffix(archr_sample)]
missing_healthy <- setdiff(HEALTHY_SAMPLES, unique(meta$archr_sample_bare))
if (length(missing_healthy) > 0) {
  warning(
    "HEALTHY_SAMPLES lists sample(s) not found in your metadata at all: ",
    paste(missing_healthy, collapse = ", "),
    " -- double-check the ArchR `Sample` spelling if you expected them."
  )
}
meta <- meta[archr_sample_bare %in% HEALTHY_SAMPLES]
if (nrow(meta) == 0) {
  stop(
    "No cells left after restricting to HEALTHY_SAMPLES. Every distinct ",
    "`Sample` value actually present in your metadata (post-PassQC) is:\n  ",
    paste(sort(unique(fread(meta_path)[PassQC == 1]$Sample)), collapse = "\n  "),
    "\nUpdate HEALTHY_SAMPLES with the six that correspond to your healthy ",
    "donors, using their exact spelling from this list (or its bare form, ",
    "stripping any _fragments.<ext>.gz suffix)."
  )
}

# ---- 2b. map each observed ArchR `Sample` value to its GEO fragments file --
crosswalk <- fread(crosswalk_path)
archr_samples <- unique(meta$archr_sample)
hit <- match(strip_suffix(archr_samples), crosswalk$archr_sample_guess)
sample_map <- setNames(
  file.path(fragments_dir, crosswalk$fragment_filename[hit]),
  archr_samples
)
sample_map <- sample_map[!is.na(hit)]  # unresolved entries are reported below, not silently NA-mapped

missing_samples <- setdiff(archr_samples, names(sample_map))
if (length(missing_samples) > 0) {
  stop(
    "No fragments file mapped for sample(s): ",
    paste(missing_samples, collapse = ", "),
    "\nThese don't match any `archr_sample_guess` in geo_sample_crosswalk.csv ",
    "(after stripping a trailing _fragments.<ext>.gz) -- likely a HIV/",
    "organophosphate/DU19 donor whose ArchR `Sample` naming wasn't confirmed; ",
    "check its row in the crosswalk CSV and add/fix it there."
  )
}

# Matching a Sample value to a crosswalk row doesn't guarantee the resulting
# path actually exists on disk -- `fragments_dir` may be wrong, or the local
# file may have been downloaded/renamed without the "GSMxxxxxxx_" prefix that
# `fragment_filename` carries. Fail fast here with the full list of
# unresolved paths rather than dying inside the per-sample loop below on
# whichever file happens to come first.
missing_files <- sample_map[!file.exists(sample_map)]
if (length(missing_files) > 0) {
  stop(
    "sample_map resolves to fragments files that don't exist on disk:\n",
    paste(sprintf("  %-25s -> %s", names(missing_files), missing_files), collapse = "\n"),
    "\nCheck that `fragments_dir` (currently \"", fragments_dir, "\") is where the ",
    "GEO fragment .bed.gz files actually live, and that they still carry the ",
    "\"GSMxxxxxxx_\" prefix from geo_sample_crosswalk.csv's `fragment_filename` ",
    "column -- if you renamed them on download, either rename them back or ",
    "adjust `fragment_filename` in the crosswalk CSV to match."
  )
}

# ---- 3. gene annotation (shared across all samples) ------------------------
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevelsStyle(annotations) <- "UCSC"
genome(annotations) <- genome_build

# ---- 4. genome-wide 5kb bins as the feature space (matches this repo's own
#         convention -- see `create_matrix_bins` / `5kb_bin_genome.sh`, which
#         do `samtools idxstats` -> `bedtools makewindows -w 5000` -> `fragtk
#         matrix -f fragments.tsv.gz -b windows.bed`). Signac's FeatureMatrix()
#         takes any GRanges of features, so tiling the genome into 5kb bins
#         and using that in place of a peak set is the direct equivalent --
#         no macs2 call needed.
BIN_WIDTH <- 5000

chrom_sizes <- seqlengths(BSgenome.Hsapiens.UCSC.hg38)   # swap genome if not hg38
bins <- tileGenome(
  chrom_sizes,
  tilewidth = BIN_WIDTH,
  cut.last.tile.in.chrom = TRUE
)
bins <- keepStandardChromosomes(bins, pruning.mode = "coarse")
seqlevelsStyle(bins) <- "UCSC"
peaks <- bins   # kept the `peaks` name below so the rest of the script (FeatureMatrix, etc.) is unchanged

# ---- 5. build one ChromatinAssay/Seurat object per ArchR sample -----------

obj_list <- list()

for (s in unique(meta$archr_sample)) {
  meta_s <- meta[archr_sample == s]

  # These GEO ArchR-exported fragment files keep the FULL "{Sample}#{barcode}"
  # composite in column 4 (e.g. "ATAC_HIP045_fragments.tsv.gz#AAACGAA...-1"),
  # NOT the bare 10x barcode -- and that composite is exactly cellId_archr.
  # (Matching on meta_s$barcode fails with "Not all cells requested could be
  # found in the fragment file".) cellId_archr already carries the sample
  # prefix, so it is unique across samples and needs no post-hoc renaming.
  cells_s <- meta_s$cellId_archr

  frag_s <- CreateFragmentObject(
    path  = sample_map[[s]],
    cells = cells_s
  )

  counts_s <- FeatureMatrix(
    fragments = frag_s,
    features  = peaks,
    cells     = cells_s
  )

  assay_s <- CreateChromatinAssay(
    counts     = counts_s,
    fragments  = frag_s,
    annotation = annotations,
    genome     = genome_build
  )

  seu_s <- CreateSeuratObject(counts = assay_s, assay = "ATAC")

  # colnames(seu_s) are already cellId_archr (== fragment-file barcodes, unique
  # across samples), so key the metadata on cellId_archr directly -- no
  # RenameCells needed, and AddMetaData lines up unambiguously.
  meta_s_df <- as.data.frame(meta_s)
  rownames(meta_s_df) <- meta_s_df$cellId_archr
  seu_s <- AddMetaData(seu_s, metadata = meta_s_df)

  obj_list[[s]] <- seu_s
}

# ---- 6. merge all samples into one combined object -------------------------
combined <- if (length(obj_list) == 1) {
  obj_list[[1]]
} else {
  merge(obj_list[[1]], y = obj_list[-1])
}

# merge() does not carry Fragment objects over onto the combined assay, so
# they must be re-attached. Pull them straight from the pre-merge per-sample
# objects rather than rebuilding from scratch -- each Fragment's cell mapping
# already uses cellId_archr (== the raw file barcode), so this preserves it.
Fragments(combined) <- NULL
for (s in names(obj_list)) {
  Fragments(combined[["ATAC"]]) <- Fragments(obj_list[[s]])[[1]]
}

# ---- 7. carry over the ArchR UMAP + clustering as-is (no need to recompute) -
umap_coords <- as.matrix(combined[[c("UMAP_1", "UMAP_2")]])
colnames(umap_coords) <- paste0("archrUMAP_", seq_len(ncol(umap_coords)))  # DimReduc requires colnames == key + index
combined[["archr_umap"]] <- CreateDimReducObject(
  embeddings = umap_coords,
  key        = "archrUMAP_",
  assay      = "ATAC"
)
Idents(combined) <- combined$ClusterCellTypes

saveRDS(combined, "combined_signac_obj.rds")
message("Done: ", ncol(combined), " cells x ", nrow(combined), " 5kb bins -> combined_signac_obj.rds")

```


```{r umaps, fig.width=12, fig.height=6}
nm  <- "ATAC Gündüz et al., 2025, BioarXiv"                       # pick any name from names(merged.ls)
obj <- merged.ls[[nm]]

p_clu <- DimPlot( merged.ls$`PBMC|ATAC`, reduction = "umap_harmony", label = TRUE, group.by = "predicted.id_ATLAS_ATAC") + NoLegend() + ggtitle(nm)
p_smp <- DimPlot(merged.ls$`PBMC|H3K27me3`, reduction = "umap_harmony", group.by = "sample")  + ggtitle("sample")
p_pat <- DimPlot(obj, reduction = "umap_harmony", group.by = "patient") + ggtitle("patient")
ggarrange(p_clu, p_smp, p_pat, ncol = 3)

p_smp
```

Gene activty from both 


```{r}

# get gene annotations from Ensembl
annotations <- GetGRangesFromEnsDb(ensdb = genome_ann)

# change to UCSC style since the data was mapped to mm10
seqlevelsStyle(annotations) <- 'UCSC'

# add the gene information to the object
Annotation(merged.ls$`PBMC|ATAC`) <- annotations
Annotation(combined) <- annotations

```


```{r gene_activity}

# Calculate gene activities
gene.activities <- GeneActivity(merged.ls$`PBMC|ATAC`)
gene.activities <- GeneActivity(combined)
# Add the gene activity matrix to the Seurat object as a new assay
merged.ls$`PBMC|ATAC`[['RNA']] <- CreateAssayObject(counts = gene.activities)
combined[['RNA']] <- CreateAssayObject(counts = gene.activities)
# Set RNA as default assay
DefaultAssay(merged.ls$`PBMC|ATAC`) <- 'RNA'
DefaultAssay(combined) <- 'RNA'
# normalize RNA data
merged.ls$`PBMC|ATAC` <- NormalizeData(merged.ls$`PBMC|ATAC`) |>
            FindVariableFeatures() |>
            ScaleData()

combined <- NormalizeData(combined) |>
            FindVariableFeatures() |>
            ScaleData()

```

label transfer with ATAC



```{r}

feats   <- intersect(VariableFeatures(merged.ls$`PBMC|ATAC`@assays$RNA), rownames(combined))
 
  anchors <- FindTransferAnchors(reference = combined, query = merged.ls$`PBMC|ATAC`,
               reference.assay = "RNA", query.assay = "RNA",
               reduction = "cca", dims = 1:50, features = feats)

merged.ls$`PBMC|ATAC`<- MapQuery(
  anchorset =anchors, 
  reference = combined,
  query =   merged.ls$`PBMC|ATAC`,
  refdata = "ClusterCellTypes")
 
#rename predictions columns 
cn <- colnames(merged.ls$`PBMC|ATAC`@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_ATLAS_ATAC"
cn[cn == "predicted.id.score"] <- "predicted.id.score_ATLAS_ATAC"
colnames(merged.ls$`PBMC|ATAC`@meta.data) <- cn

#compacting cell annotation

merged.ls$`PBMC|ATAC`@meta.data$BroadCellType <- case_when(
  merged.ls$`PBMC|ATAC`@meta.data$predicted.id_ATLAS_ATAC %in% c("T_naive","T_mem_CD4","T_mem_CD8","T_mix") ~ "T_cell",
  merged.ls$`PBMC|ATAC`@meta.data$predicted.id_ATLAS_ATAC %in% c("B_naive","B_mem") ~ "B_cell",
  merged.ls$`PBMC|ATAC`@meta.data$predicted.id_ATLAS_ATAC %in% c("Plasma") ~ "Plasmacell",
  merged.ls$`PBMC|ATAC`@meta.data$predicted.id_ATLAS_ATAC %in% c("NK_CD16") ~ "NK_cell",
  merged.ls$`PBMC|ATAC`@meta.data$predicted.id_ATLAS_ATAC %in% c("Mono_CD14","Mono_CD16","Mono_other") ~ "Monocytes",
   TRUE                                     ~ NA_character_
)



saveRDS(merged.ls,"/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects/merged_PBMC.rds")
```
finally assign the final labels to all the cells that are shared between the 3 modalities

```{r}

cell_labels<-merged.ls$`PBMC|ATAC`@meta.data %>% dplyr::select(barcode,predicted.id_ATLAS_ATAC,predicted.id.score_ATLAS_ATAC)

merged.ls$`PBMC|H3K27ac`@meta.data<- full_join(merged.ls$`PBMC|H3K27ac`@meta.data,cell_labels)

cell_labels2<-merged.ls$`PBMC|H3K27ac`@meta.data %>% dplyr::select(barcode,predicted.id_ATLAS_ATAC,predicted.id.score_ATLAS_ATAC) %>% na.omit() %>% unique()

merged.ls$`PBMC|H3K27me3`@meta.data<- full_join(merged.ls$`PBMC|H3K27me3`@meta.data,cell_labels2)

# First step we have to extract all the barcodes that common across 4 modalities

    x <- list(name_mod1=colnames(merged.ls$`PBMC|ATAC`),
              name_mod2=colnames(merged.ls$`PBMC|H3K27ac`),
              name_mod3=colnames(merged.ls$`PBMC|H3K27me3`))
barcode_list<-Reduce(intersect, x)
# Identify the cell barcodes that you want to keep
combined.obj.ls_epi_mod<-list()
modalities<-c("PBMC|ATAC","PBMC|H3K27ac","PBMC|H3K27me3")
barcodes_to_keep <- barcode_list
for (mod in modalities){
# Use the subset function to filter the Seurat object
merged.ls[[mod]] <- subset(combined.obj.ls[[paste0(mod)]], cells = barcodes_to_keep)

# Print or further analyze your filtered Seurat object

}


saveRDS(merged.ls,"/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects/merged_PBMC.rds")

```
WNN

```{r}
DefaultAssay(merged.ls$`PBMC|ATAC`) <- "bins"
#first change colnames in the object 

merged.ls$`PBMC|ATAC`@meta.data$barcode<-paste0(merged.ls$`PBMC|ATAC`@meta.data$sample,"_",merged.ls$`PBMC|ATAC`@meta.data$barcode)
merged.ls$`PBMC|H3K27ac`@meta.data$barcode<-paste0(merged.ls$`PBMC|H3K27ac`@meta.data$sample,"_",merged.ls$`PBMC|H3K27ac`@meta.data$barcode)
merged.ls$`PBMC|H3K27me3`@meta.data$barcode<-paste0(merged.ls$`PBMC|H3K27me3`@meta.data$sample,"_",merged.ls$`PBMC|H3K27me3`@meta.data$barcode)

#change cell barcodes

merged.ls$`PBMC|ATAC`<-RenameCells(merged.ls$`PBMC|ATAC`,new.names = merged.ls$`PBMC|ATAC`@meta.data$barcode)
merged.ls$`PBMC|H3K27ac`<-RenameCells(merged.ls$`PBMC|H3K27ac`,new.names = merged.ls$`PBMC|H3K27ac`@meta.data$barcode)
merged.ls$`PBMC|H3K27me3`<-RenameCells(merged.ls$`PBMC|H3K27me3`,new.names = merged.ls$`PBMC|H3K27me3`@meta.data$barcode)

#change assay name 

merged.ls$`PBMC|ATAC`<- RenameAssays(object = merged.ls$`PBMC|ATAC`, bins = "bins_ATAC")
merged.ls$`PBMC|H3K27ac`<- RenameAssays(object = merged.ls$`PBMC|H3K27ac`, bins = "bins_H3K27ac")
merged.ls$`PBMC|H3K27me3`<- RenameAssays(object = merged.ls$`PBMC|H3K27me3`, bins = "bins_H3K27me3")

#extract assay and metadata and create a unique object

meta_ATAC <- merged.ls$`PBMC|ATAC`@meta.data
meta2_H3K27ac <- merged.ls$`PBMC|H3K27ac`@meta.data
meta3_H3K27me3 <- merged.ls$`PBMC|H3K27me3`@meta.data

# Add suffixes to avoid name clashes
colnames(meta_ATAC) <- paste0(colnames(meta_ATAC), "_ATAC")
colnames(meta2_H3K27ac) <- paste0(colnames(meta2_H3K27ac), "_H3K27ac")
colnames(meta2_H3K27ac) <- paste0(colnames(meta2_H3K27ac), "_H3K27me3")

meta_all <- cbind(meta_ATAC, meta2_H3K27ac[rownames(meta_ATAC), ], meta3_H3K27me3[rownames(meta_ATAC), ]) %>% na.omit()


ATAC  <- merged.ls$`PBMC|ATAC`@assays$bins_ATAC
keep <- intersect(Cells(ATAC), rownames(meta_all))
length(keep)  
ATAC<- subset(ATAC, cells = keep)
H3K27ac <-  merged.ls$`PBMC|H3K27ac`@assays$bins_H3K27ac
H3K27ac<-subset(H3K27ac, cells = keep)
H3K27me3  <- merged.ls$`PBMC|H3K27me3`@assays$bins_H3K27me3
H3K27me3<-subset(H3K27me3, cells = keep)


PBMCs_nanoCTAR <- CreateSeuratObject(  counts    = ATAC,
  assay     = "bins_ATAC",
  meta.data = meta_all)

PBMCs_nanoCTAR[["bins_H3K27ac"]]  <- H3K27ac
PBMCs_nanoCTAR[["bins_H3K27me3"]]  <- H3K27me3

DefaultAssay(MS_MultiOmics) <- "RNA"


```

rerun dimensionality reduction and clustering using the cells encompassing the 3 modalities

```{r}
marks <- c("ATAC", "H3K27ac", "H3K27me3")
pc    <- 10

for (mk in marks) {
  assay <- paste0("bins_", mk)
  lsi   <- paste0("lsi_",  mk)
  snn   <- paste0(assay, "_snn")
  message("=== ", assay, " ===")

  DefaultAssay(PBMCs_nanoCTAR) <- assay

  PBMCs_nanoCTAR <- RunTFIDF(PBMCs_nanoCTAR)
  PBMCs_nanoCTAR <- FindTopFeatures(PBMCs_nanoCTAR, min.cutoff = 5)
  PBMCs_nanoCTAR <- RunSVD(
    PBMCs_nanoCTAR,
    reduction.name = lsi,
    reduction.key  = paste0("LSI", mk, "_")
  )

  PBMCs_nanoCTAR <- RunUMAP(
    PBMCs_nanoCTAR,
    reduction      = lsi,
    dims           = 2:pc,
    reduction.name = paste0("umap_", mk),
    reduction.key  = paste0("UMAP", mk, "_")
  )

  PBMCs_nanoCTAR <- FindNeighbors(
    PBMCs_nanoCTAR,
    reduction  = lsi,
    dims       = 2:pc,
    graph.name = c(paste0(assay, "_nn"), snn)
  )
  PBMCs_nanoCTAR <- FindClusters(
    PBMCs_nanoCTAR,
    graph.name = snn,
    algorithm  = 3,
    resolution = 0.6,
    verbose    = FALSE
  )

  # seurat_clusters is overwritten each pass -- keep a per-modality copy
  PBMCs_nanoCTAR[[paste0("clusters_", mk)]] <- Idents(PBMCs_nanoCTAR)
}
```


Harmony


```{r}
marks <- c("ATAC", "H3K27ac", "H3K27me3")
pc    <- 10
lsi_dims  <- 2:pc            # skip LSI_1 (depth-correlated)
harm_dims <- 1:(pc - 1)      # Harmony returns length(lsi_dims) components

stopifnot(length(unique(PBMCs_nanoCTAR$patient)) > 1)

for (mk in marks) {
  assay <- paste0("bins_", mk)
  lsi   <- paste0("lsi_", mk)
  harm  <- paste0("harmony_lsi_", mk)
  snn   <- paste0(assay, "_snn")
  hsnn  <- paste0(assay, "_harmony_snn")
  message("=== ", assay, " ===")

  DefaultAssay(PBMCs_nanoCTAR) <- assay

  ## ---- Harmony-integrated ----
  PBMCs_nanoCTAR <- RunHarmony(
    PBMCs_nanoCTAR, group.by.vars = "patient",
    reduction.use  = lsi,
    dims.use       = lsi_dims,
    reduction.save = harm,
    project.dim    = FALSE
  )
  PBMCs_nanoCTAR <- RunUMAP(
    PBMCs_nanoCTAR, reduction = harm, dims = harm_dims,
    reduction.name = paste0("umap_harmony_", mk),
    reduction.key  = paste0("UMAPh", mk, "_")
  )
  PBMCs_nanoCTAR <- FindNeighbors(
    PBMCs_nanoCTAR, reduction = harm, dims = harm_dims,
    graph.name = c(paste0(assay, "_harmony_nn"), hsnn)
  )
  PBMCs_nanoCTAR <- FindClusters(
    PBMCs_nanoCTAR, graph.name = hsnn,
    algorithm = 3, resolution = 0.6, verbose = FALSE
  )
  PBMCs_nanoCTAR[[paste0("harmony_clusters_", mk)]] <- Idents(PBMCs_nanoCTAR)
}
```


WNN

```{r}
DefaultAssay(PBMCs_nanoCTAR) <- "bins_ATAC"

PBMCs_nanoCTAR <- FindMultiModalNeighbors(
 PBMCs_nanoCTAR,
  reduction.list = list("lsi_ATAC", "lsi_H3K27ac", "lsi_H3K27me3"),
  dims.list      = list(1:10,      1:10,          1:10),
  modality.weight.name = "ATAC.weight"  # name for stored modality weights
)

PBMCs_nanoCTAR <- FindMultiModalNeighbors(
 PBMCs_nanoCTAR,
  reduction.list = list("harmony_lsi_ATAC", "harmony_lsi_H3K27ac", "harmony_lsi_H3K27me3"),
  dims.list      = list(1:9,      1:9,          1:9),
  modality.weight.name = "ATAC.weight",  # name for stored modality weights
   knn.graph.name = "wknn_harmony",
  snn.graph.name = "wsnn_harmony",
  weighted.nn.name = "weighted.nn_harmony",
)


```

Clustering and UMAP based on the WNN


```{r}
# Cluster on the WNN SNN graph
PBMCs_nanoCTAR   <- FindClusters(
  PBMCs_nanoCTAR,
  graph.name = "wsnn",
  algorithm = 3,        # Leiden
  resolution = 0.5
)

# UMAP using WNN neighbors

PBMCs_nanoCTAR  <-RunUMAP(PBMCs_nanoCTAR, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")


```

```{r,fig.height=8, fig.width=14}

library("randomcoloR")
set.seed(1)
cluster.cols <- distinctColorPalette(6) 


p1<-DimPlot(PBMCs_nanoCTAR,reduction = "wnn.umap_harmony",group.by = "BroadCellType_ATAC",cols = cluster.cols)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("WNN all modalities") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

p2<-DimPlot(PBMCs_nanoCTAR,reduction = "wnn.umap",group.by = "BroadCellType_ATAC",cols = cluster.cols)+
  labs(x = "UMAP 1", y = "UMAP 2") +
  ggtitle("WNN all modalities") +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20),
    legend.text = element_text(size = 16),         # Legend labels
    legend.title = element_text(size = 24),  # Legend title
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )


p1
p2
```


```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

modality<-c("ATAC","H3K27ac","H3K27me3")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/PBMCs_overlap_metadata.xlsx"

  md <- PBMCs_nanoCTAR@meta.data
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))

saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/PBMCs_UMAP_overlap_coordinates_harmony.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in modality) {
  md <- PBMCs_nanoCTAR@reductions[[paste0("umap_harmony_",mk)]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/PBMCs_overlap_UMAP_coordinates.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in modality) {
  md <- PBMCs_nanoCTAR@reductions[[paste0("umap_",mk)]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/PBMCs_overlap_WNN_coordinates.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
  md <- PBMCs_nanoCTAR@reductions[["wnn.umap"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb,sheetName = "wnn")
  writeData(wb,sheet = "wnn",df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))

saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")


obj_dir  <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/PBMCs_overlap_WNN_harmony_coordinates.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
  md <- PBMCs_nanoCTAR@reductions[["wnn.umap_harmony"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb,sheetName = "wnn")
  writeData(wb,sheet = "wnn",df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))

saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

saveRDS(PBMCs_nanoCTAR,"/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS/first_eval/objects/PBMC_multimodal.rds")
```


```{r session}
sessionInfo()
```
