---
title: "First evaluation — interactive"
author: "Mattia Zaghi"
output:
  html_document:
    toc: true
    toc_float: true
---

Interactive companion to `first_eval_MGI.Rmd`. Unlike the report (which is
parametrised and meant to be rendered by snakemake), this one **defines its
config in a plain chunk** so you can run it **chunk-by-chunk in RStudio**
(Ctrl/Cmd+Enter) and continue the analysis. It **loads the objects the report
already saved** — no rebuilding.

Needs an R with the analysis packages (Signac/Seurat/...): your usual RStudio R,
or `conda activate nanoctarna-analysis; rstudio`.

```{r setup, message=FALSE, warning=FALSE}
# ---- libraries + helper functions + genome --------------------------------
suppressPackageStartupMessages({
  library(Signac); library(Seurat); library(GenomicRanges)
  library(stringr); library(dplyr); library(ggplot2); library(ggpubr)
  library(yaml); library(ggVennDiagram); library(ComplexUpset)
  library(EnsDb.Hsapiens.v86); library(BSgenome.Hsapiens.UCSC.hg38)
})
repodir <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction"
source(file.path(repodir, "scripts", "functions_scCT.R"))       # plotCounts, DepthCorMulMod, ...
source(file.path(repodir, "scripts", "functions_firstEval.R"))  # config-driven first-eval helpers

genome     <- "hg38"
genome_ann <- EnsDb.Hsapiens.v86
bs_genome  <- BSgenome.Hsapiens.UCSC.hg38
```

## Config (edit if your run is elsewhere)

`params` is a plain list here (not YAML), so every chunk that uses `params$...`
— including chunks copied from `first_eval_MGI.Rmd` — works when run interactively.

```{r config}
wd         <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/"   # analysis root
result_dir <- file.path(wd, "first_eval")
obj_dir    <- file.path(result_dir, "objects")

params <- list(
  wd            = paste0(wd, "/"),
  binsize       = 5000,
  filter_passed = TRUE,
  dims          = 30,
  resolution    = 0.6,
  workers       = 4,
  group_pattern = "^(MS)",
  fig_dir       = file.path(result_dir, "figures"),
  fig_scale     = 1.25
)
dir.create(params$fig_dir, recursive = TRUE, showWarnings = FALSE)
fig_scale <- params$fig_scale
SP <- function(p, nm, w = 8, h = 6)                     # same figure saver as the report
  save_png(p, nm, params$fig_dir, w * fig_scale, h * fig_scale)
```

## Load what the report already built

```{r load-objects}
load_if <- function(f) if (file.exists(f)) readRDS(f) else {
  message("not found (run the report first): ", f); NULL }

obj.ls    <- load_if(file.path(obj_dir, "obj.ls_bins.rds"))            # list keyed by experiment
#merged.ls <- load_if(file.path(obj_dir, "merged.ls_group_mark.rds"))  # list keyed by "group|mark"
#cell_df   <- load_if(file.path(obj_dir, "qc_cell_df.rds"))            # per-cell QC table
# also available individually (per group), and the raw 5kb matrices:
#   merged_MS   <- load_if(file.path(obj_dir, "merged_MS.rds"))       # list(H3K27ac=, H3K27me3=)
#   counts.ls   <- load_if(file.path(obj_dir, paste0("counts.ls_bins_", params$binsize, ".rds")))

cat("obj.ls   :", if (is.null(obj.ls))    "-" else paste(length(obj.ls), "experiments"), "\n",
    "merged.ls:", if (is.null(merged.ls)) "-" else paste(names(merged.ls), collapse = ", "), "\n",
    "cell_df  :", if (is.null(cell_df))   "-" else paste(nrow(cell_df), "cells"), "\n")
```

## Doublet removal (AMULET) -- must run before merge/cluster, on obj.ls

`obj.ls` still has one Signac object per experiment with native (unprefixed)
barcodes -- exactly what AMULET needs, since it scores multiplets from a
single experiment's `fragments_raw.tsv.gz`. We flag+drop doublets here, then
rebuild `merged.ls` from the cleaned `obj.ls` (cheap: `merge_by_group_mark`
just re-merges already-computed objects, no matrix rebuild).

```{r doublets_amulet}
amulet_dir <- "/home/mattia/AMULET"                                   # AMULET.py location
chrlist    <-  file.path(amulet_dir, "ref/chrName.txt")        # TODO: add to config/ (hg38)
blacklist  <- file.path(amulet_dir, "ref/hg38.blacklist.bed") # TODO: add to config/ (hg38)

# AMULET needs pandas/numpy/pysam -- neither nanoscope_general nor
# nanoscope_debarcode ship pandas, so it gets its own tiny env
# (envs/nanoscope_amulet.yaml): `conda env create -f envs/nanoscope_amulet.yaml`.
# system() spawns a plain subshell that does NOT re-source your `conda activate`,
# so point it at the env's python by absolute path rather than relying on
# whatever "python" happens to be on PATH when this is run from RStudio.
amulet_python <- "/home/mattia/miniconda3_n//envs/nanoscope_amulet/bin/python3"  # adjust to your conda prefix
stopifnot("amulet_python not found -- fix the path or `conda env create -f envs/nanoscope_amulet.yaml`" =
            file.exists(amulet_python))

# exp_id -> fragments file, without assuming barcode format. Splits on the
# FIRST "." (sample never contains one; modality might in principle, but
# doesn't in this config) and globs <wd>/<sample>/<modality>_*/... for the
# cellranger outs dir, so it matches "H3K27ac_MS058", "H3K27ac_CCTATCCT", etc.
find_fragments <- function(exp_id, wd) {
  sample   <- sub("\\..*$", "", exp_id)
  modality <- sub("^[^.]+\\.", "", exp_id)
  exp_dir  <- Sys.glob(file.path(wd, sample, paste0(modality, "_*")))
  if (length(exp_dir) == 0) return(NA_character_)
  cand <- Sys.glob(file.path(exp_dir, "cellranger", "outs",
                             c("fragments_raw.tsv.gz", "fragments.tsv.gz")))
  if (length(cand) == 0) return(NA_character_)
  cand[1]
}

# AMULET is a two-step CLI, not one call (there is no --forcesorted flag on
# either script -- that's what the earlier run errored on):
#   1) FragmentFileOverlapCounter.py <fragments> <bcmap.csv> <chrlist> <outdir>
#      -> writes Overlaps.txt / OverlapSummary.txt into <outdir>. <bcmap.csv>
#      must be a csv *with a header row*, columns "barcode" and
#      "is__cell_barcode" (rows kept where == 1) -- not the one-barcode-
#      per-line file this chunk used to write.
#   2) AMULET.py --rfilter <blacklist.bed> <Overlaps.txt> <OverlapSummary.txt> <outdir2>
#      -> the repeat blacklist is only ever an AMULET.py argument
#      (--rfilter); it does not go into step 1 at all.
flag_doublets_amulet <- function(exp_id, obj, wd, qval_cutoff = 0.05) {
  frag <- find_fragments(exp_id, wd)
  if (is.na(frag)) { message("no fragments for ", exp_id, ", skipping"); return(obj) }

  counter_dir <- file.path(obj_dir, "amulet", exp_id, "overlaps")
  amulet_out  <- file.path(obj_dir, "amulet", exp_id, "multiplets")
  dir.create(counter_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(amulet_out,  recursive = TRUE, showWarnings = FALSE)

  # BCMAP: header "barcode,is__cell_barcode" -- score only cells that passed
  # pick_cells.R, so every row here is a valid cell (is__cell_barcode = 1)
  bc_file <- file.path(counter_dir, "singlecell.csv")
  write.csv(data.frame(barcode = colnames(obj), is__cell_barcode = 1L,
                        check.names = FALSE),
            bc_file, row.names = FALSE, quote = FALSE)

  # step 1: per-fragment overlap counting (no --rfilter here)
  system(sprintf("%s %s/FragmentFileOverlapCounter.py %s %s %s %s",
                  amulet_python, amulet_dir, frag, bc_file, chrlist, counter_dir))

  overlaps_path <- file.path(counter_dir, "Overlaps.txt")
  summary_path  <- file.path(counter_dir, "OverlapSummary.txt")
  if (!file.exists(overlaps_path) || !file.exists(summary_path)) {
    message("FragmentFileOverlapCounter produced no output for ", exp_id); return(obj)
  }

  # step 2: multiplet probability estimation; blacklist goes here via --rfilter
  system(sprintf("%s %s/AMULET.py --rfilter %s %s %s %s",
                  amulet_python, amulet_dir, blacklist, overlaps_path, summary_path, amulet_out))

  probs_path <- file.path(amulet_out, "MultipletProbabilities.txt")
  if (!file.exists(probs_path)) { message("AMULET produced no output for ", exp_id); return(obj) }

  # AMULET.py writes columns cell_id, barcode, p-value, q-value and already
  # BH-corrects q-value itself -- no need to re-run p.adjust. Match by
  # barcode explicitly rather than assuming row order/rownames.
  probs <- read.delim(probs_path, check.names = FALSE)
  qval  <- setNames(probs[["q-value"]], probs[["barcode"]])[colnames(obj)]
  obj$amulet_qval       <- qval
  obj$is_doublet_amulet <- qval < qval_cutoff
  obj
}

obj.ls <- Map(flag_doublets_amulet, names(obj.ls), obj.ls, MoreArgs = list(wd = params$wd))
saveRDS(obj.ls, file.path(obj_dir, "obj.ls_bins_Amulet.rds"))
n_scored   <- sum(sapply(obj.ls, function(o) !is.null(o$is_doublet_amulet)))
n_doublets <- sum(sapply(obj.ls, function(o) if (is.null(o$is_doublet_amulet)) 0 else sum(o$is_doublet_amulet, na.rm = TRUE)))
cat(n_doublets, "multiplets flagged across", n_scored, "of", length(obj.ls), "experiments",
    "(the rest had no fragments file found -- check find_fragments() against your layout)\n")

# drop flagged cells from obj.ls (native names) and from merged.ls (which may
# have prefixed names "<exp_id>_<barcode>" only for marks merged from >1 exp)
for (id in names(obj.ls)) {
  o <- obj.ls[[id]]
  if (is.null(o$is_doublet_amulet)) next
  bad <- colnames(o)[o$is_doublet_amulet]
  if (length(bad) == 0) next
  obj.ls[[id]] <- subset(o, cells = colnames(o)[!o$is_doublet_amulet])
  for (nm in names(merged.ls)) {
    m    <- merged.ls[[nm]]
    drop <- intersect(colnames(m), c(bad, paste0(id, "_", bad)))
    if (length(drop) > 0) merged.ls[[nm]] <- subset(m, cells = setdiff(colnames(m), drop))
  }
}
```
create a merged MS object

```{r merge_modalities}

merged.ls <- merge_by_group_mark(obj.ls, exp_df, group_pattern = params$group_pattern)
cat("Joint groups (type x mark):", paste(names(merged.ls), collapse = ", "), "\n")

grp_names <- names(merged.ls)
merged.ls <- .par_lapply(grp_names, function(nm) {
  tryCatch(
    normalize_dr(merged.ls[[nm]], min_cutoff = 5),
    error = function(e) {
      message("normalize_dr failed for '", nm, "': ", conditionMessage(e))
      merged.ls[[nm]]
    }
  )
})
names(merged.ls) <- grp_names
```




```{r batch_harmony}

suppressPackageStartupMessages(library(harmony))

# 1) technology, from sample
merged.ls$`H3K27ac`@meta.data$technology <- case_when(
  merged.ls$`H3K27ac`@meta.data$sample %in% c("MS3BL1_CT", "MS3BL2_CT") ~ "GEM-X",
  merged.ls$`H3K27ac`@meta.data$sample  == "MS549_CT"                ~ "Next-GEM",
  TRUE                                     ~ NA_character_
)

merged.ls$`H3K27me3`$technology <- case_when(
  merged.ls$`H3K27me3`@meta.data$sample %in% c("MS3BL1_CT", "MS3BL2_CT") ~ "GEM-X",
  merged.ls$`H3K27me3`@meta.data$sample  == "MS549_CT"                ~ "Next-GEM",
  TRUE                                     ~ NA_character_
)


# 2) protocol, from sample AND the technology just derived above
merged.ls$`H3K27ac`@meta.data$protocol <- case_when(
  merged.ls$`H3K27ac`@meta.data$patient %in% c("MS058CT1", "MS058CT2", "MS381CT1","MS381CT2","MS549_CT") ~ "fresh",
  merged.ls$`H3K27ac`@meta.data$patient == "MS549CT" ~ "frozen",
  TRUE                                                                        ~ NA_character_
)

merged.ls$`H3K27me3`@meta.data$protocol <- case_when(
  merged.ls$`H3K27me3`@meta.data$patient %in% c("MS058CT1", "MS058CT2", "MS381CT1","MS381CT2","MS549_CT") ~  "fresh" ,
  merged.ls$`H3K27me3`@meta.data$patient == "MS549CT"~"frozen",
  TRUE                                                                        ~ NA_character_
)


merged.ls <- lapply(merged.ls, function(obj) {
  #obj <- RunTFIDF(obj) > FindTopFeatures(min.cutoff = 5) > RunSVD()
  obj <- RunHarmony(obj, group.by.vars = c("technology","protocol"),
                    reduction.use = "lsi", reduction.save = "harmony_lsi",
                    project.dim = FALSE)
  obj <- RunUMAP(obj, reduction = "harmony_lsi", di = 2:params$dims, reduction.name = "umap_harmony")
  obj <- FindNeighbors(obj, reduction = "harmony_lsi", di = 2:params$dims)
  obj <- FindClusters(obj, resolution = params$resolution, algorithm = 3, verbose = FALSE,cluster.name="harmony_clusters",graph.name = "bins_snn")
  obj
})
```


```{r umaps, fig.width=12, fig.height=6}
nm  <- "MS|H3K27ac"                       # pick any name from names(merged.ls)
obj <- merged.ls[[nm]]

p_clu <- DimPlot(obj, reduction = "umap_harmony", label = TRUE) + NoLegend() + ggtitle(nm)
p_smp <- DimPlot(obj, reduction = "umap_harmony", group.by = "sample")  + ggtitle("sample")
p_pat <- DimPlot(obj, reduction = "umap_harmony", group.by = "patient") + ggtitle("patient")
ggarrange(p_clu, p_smp, p_pat, ncol = 3)

```

```{r recluster}
# re-cluster / re-embed with different parameters and inspect
obj <- FindClusters(obj, resolution = 0.4, algorithm = 3, verbose = FALSE)
DimPlot(obj, reduction = "umap", label = TRUE) + NoLegend()
```

## Continue: QC of any metric (from the saved cell_df)

```{r qc}
# metrics available: logUMI, nCount_bins, passed_filters, reads, dup_rate, FRiP, TSS_enrichment
qc_metric_violin(cell_df, "nCount_bins", ylab = "unique fragments (bins)", log_y = TRUE)
# per group:
# qc_metric_violin(cell_df, "TSS_enrichment", ylab = "TSS enrichment", group = "MS")
```

## (Optional) rebuild instead of loading

Only if you changed something upstream. Reuses the cached 5kb matrices, so it's fast.

```{r rebuild, eval=FALSE}
exp_df  <- discover_experiments(params$wd)
meta.ls <- load_metadata_list(exp_df, params$wd, filter_passed = params$filter_passed)
obj.ls  <- build_bins_objects(
  exp_df, meta.ls, params$wd, bs_genome = bs_genome, binsize = params$binsize,
  counts_cache = readRDS(file.path(obj_dir, paste0("counts.ls_bins_", params$binsize, ".rds"))))
```

## Create gene activity for label transfer with Mukund Dataset
### 6.1 Add gene annotations to the Seurat object
Before creating a gene activity matrix we first need to add gene annotations to the H3K27ac Seurat object.
```{r add_gene_ann }
# get gene annotations from Ensembl
annotations <- GetGRangesFromEnsDb(ensdb = genome_ann)

# change to UCSC style since the data was mapped to mm10
seqlevelsStyle(annotations) <- 'UCSC'

# add the gene information to the object
Annotation(merged.ls$`MS|H3K27ac`) <- annotations
```

```{r gene_activity}

# Calculate gene activities
gene.activities <- GeneActivity(merged.ls$`MS|H3K27ac`)

# Add the gene activity matrix to the Seurat object as a new assay
merged.ls$`MS|H3K27ac`[['RNA']] <- CreateAssayObject(counts = gene.activities)
# Set RNA as default assay
DefaultAssay(merged.ls$`MS|H3K27ac`) <- 'RNA'
# normalize RNA data
merged.ls$`MS|H3K27ac` <- NormalizeData(merged.ls$`MS|H3K27ac`) |>
            FindVariableFeatures() |>
            ScaleData()

```

```{r}
Human_H3K27ac_nanoCT <- readRDS("/date/gcb/gcb_wq/MS_project/reference_datasets/Kabbe_2026_nCT_rds/processed_data/H3K27ac_GRS_anno_EA.rds")

DefaultAssay(Human_H3K27ac_nanoCT) <- 'PROM2KBGB'
  
feats   <- intersect(VariableFeatures(MS$H3K27ac@assays$RNA), rownames(Human_H3K27ac_nanoCT))
 
anchors <- FindTransferAnchors(reference = Human_H3K27ac_nanoCT, query = MS$H3K27ac,
               reference.assay = "PROM2KBGB", query.assay = "RNA",
               reduction = "cca", dims = 1:50, features = feats)

MS$H3K27ac <- MapQuery(
  anchorset =anchors, 
  reference = Human_H3K27ac_nanoCT,
  query =   MS$H3K27ac,
  refdata = "transferred_annot")
 
#rename predictions columns 
cn <- colnames(MS$H3K27ac@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_CZI_Human"
cn[cn == "predicted.id.score"] <- "predicted.id.score_CZI_Human"
colnames(MS$H3K27ac@meta.data) <- cn


```


```{r}
Human_MS<- readRDS("/date/gcb/gcb_YD/MS_sc_project/MSref_brain_EA/ms_lesions_snRNAseq_seuratV5_WMEA_updated_fullFineLesionDown500.rds")
new_names <- sub("-ENSG[0-9]+$", "", rownames(Human_MS))
sum(duplicated(new_names))

DefaultAssay(Human_MS) <- 'RNA'


for (a in Assays(Human_MS)) {
  rn <- rownames(Human_MS[[a]])
  new_rn <- sub("-ENSG[0-9]+$", "", rn)
  if (any(duplicated(new_rn))) new_rn <- make.unique(new_rn)
  rownames(Human_MS[[a]]) <- new_rn
}


feats   <- intersect(VariableFeatures(merged.ls$`MS|H3K27ac`@assays$RNA), rownames(Human_MS))
 
anchors <- FindTransferAnchors(reference = Human_MS, query = merged.ls$`MS|H3K27ac`,
               reference.assay = "RNA", query.assay = "RNA",
               reduction = "cca", dims = 1:50, features = feats)

merged.ls$`MS|H3K27ac` <- MapQuery(
  anchorset =anchors, 
  reference = Human_MS,
  query =   merged.ls$`MS|H3K27ac`,
  refdata = "type_broad")
 
#rename predictions columns 
cn <- colnames(merged.ls$`MS|H3K27ac`@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Human_MS"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Human_MS"
colnames(merged.ls$`MS|H3K27ac`@meta.data) <- cn

```

```{r}
suppressPackageStartupMessages({ library(Seurat); library(openxlsx) })

obj_dir  <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/MS_metadata.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@meta.data
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/MS_UMAP_coordinates_harmony.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@reductions[["umap_harmony"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")

obj_dir  <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/objects"
out_xlsx <- "/date/gcb/gcb_wq/nanoCTAR_pipeline_barcode_correction/first_eval/MS_UMAP_coordinates.xlsx"


cat("marks:", paste(names(merged.ls), collapse = ", "), "\n")

wb <- createWorkbook()
for (mk in names(merged.ls)) {
  md <- merged.ls[[mk]]@reductions[["umap"]]@cell.embeddings
  df <- data.frame(cell = rownames(md), md, row.names = NULL,
                   check.names = FALSE, stringsAsFactors = FALSE)
  addWorksheet(wb, mk)
  writeData(wb, mk, df)
  cat(sprintf("  %-9s : %d cells x %d metadata columns\n", mk, nrow(df), ncol(df) - 1))
}
saveWorkbook(wb, out_xlsx, overwrite = TRUE)
cat("Wrote:", out_xlsx, "\n")
```

```{r session}
sessionInfo()
```
