---
title: "nanoCUT&Tag vs DynaTag: peak concordance and mark specificity"
author: "Mattia Zaghi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dev = "png", dpi = 150)
```

# Purpose and what changed

The previous version of this vignette asked *how much do the peak sets overlap*.
That question cannot answer the one that matters here, which is why **different
histone marks agree with each other**, most strongly in DynaTag. Three
explanations produce the same overlap numbers:

1. genuine co-occupancy (bivalent promoters, poised enhancers);
2. threshold instability — peaks flickering across the MACS q-value cutoff;
3. mark-**nonspecific** signal: untethered pA-Tn5, antibody-barcode hopping, or
   antibody cross-reactivity.

Separating them needs (a) an **external anchor** — ENCODE regions that are
exclusively one mark, so co-occupancy is excluded by construction; (b)
**normalisation for peak breadth and library depth**, so a broader or deeper
library cannot win on size; and (c) **fragment-level** rather than peak-level
counting, because peak coverage saturates once a region is called.

Four measurement bugs in the previous version are also fixed, and it is worth
knowing what each one did, because they biased the result in opposite directions:

| Old code | Problem |
|---|---|
| `length(findOverlaps(gr1,gr2))/length(union(gr1,gr2))` | counts *hit pairs*, not shared bases. One broad peak hitting 50 narrow ones contributes 50. Not bounded by 1 and not a Jaccard index. |
| no `reduce()` before set operations | `--call-summits` emits `peak_1a`/`peak_1b` at **identical** coordinates; ~19–21% of intervals in these files are duplicates, inflating every count. |
| `findOverlaps(..., type = "within")` | reports how many peaks are *contained* in the other set — a peak-width ratio, so the broad-mark comparison measures breadth, not agreement. |
| `venn_data <- list(A = seq_len(length(...)), ...)` | intersects **integer sequences**, so the Venn shows `min(n)` regardless of genomic position. Any three sets of similar size give the same picture. |

Blacklist and main-chromosome filtering were also absent; unfiltered hg38
blacklist regions are pile-up artefacts shared by *every* assay and therefore
manufacture cross-mark concordance on their own.

```{r config}
suppressPackageStartupMessages({
  library(GenomicRanges); library(rtracklayer); library(data.table)
  library(ggplot2); library(dplyr); library(tidyr); library(scales);library(GenomeInfoDb)
})

## ---- paths -----------------------------------------------------------------
## ONE root. Everything else is derived from it. Do not hand-type absolute paths
## below: six near-identical strings is exactly how "results_H3K27ac" +
## "PBMC_DynaTag_H3K27ac" lost its "/" and became a silently unreadable file
## that reported 0 fragments instead of erroring.
ROOT <- "/date/gcb/gcb_MZ/multiNanoCT/samples/PBMCs-MS"

params_ <- list(
  root         = ROOT,
  peaks_dir    = file.path(ROOT, "IGV_clusters", "peaks"),
  encode_dir   = file.path(ROOT, "Refs", "peaks"),
  blacklist    = file.path(ROOT, "Refs", "hg38-blacklist.v2.bed.gz"),
  clusters_dir = file.path(ROOT, "IGV_clusters"),

  ## out_dir  = where THIS vignette writes its own figures and exports.
  ## refs_dir = where run_specificity_all.sh wrote its cached TSVs.
  ## These are different concepts. They were previously coupled as
  ## refs_dir <- file.path(out_dir, "refs"), which silently pointed at an empty
  ## auto-created directory when the driver's OUT lived elsewhere; the symptom
  ## was "No budget_*.tsv found" even with run_fragment_analysis = FALSE.
  ## refs_dir MUST equal OUT= in run_specificity_all.sh, character for character.
  out_dir      = file.path(ROOT, "peaks_analysis"),
  refs_dir     = file.path(ROOT, "peaks_analysis"),

  ## ABSOLUTE. A relative path resolves against the .Rmd's own directory when
  ## knitting, not against your working directory. Note the spelling: the file
  ## is mark_specificity.sh (a "mark_specificty.sh" typo here fails only when a
  ## fragment section actually runs, i.e. long after setup appears to succeed).
  helper       = file.path(ROOT, "peaks_analysis", "mark_specificity.sh"),

  ## BOTH of these are required for sections 7-10. Filling `frags` alone does
  ## nothing: the flag is a separate gate, because these passes take tens of
  ## minutes per library. Once the driver has produced the TSVs, leave this
  ## FALSE and the chunks read the cache in refs_dir.
  run_fragment_analysis = FALSE,

  ## nanoscope's sort_sinto_output rule already coordinate-sorts fragments.tsv.gz,
  ## so this is normally safe and avoids an external sort of a 135M-line file.
  ## If bedtools complains about chromosome sort order, set FALSE.
  frags_presorted = TRUE
)

## library name -> antibody mark. Single source of truth for the fragment paths
## AND the mark labels used throughout, so the two cannot disagree.
lib_mark <- c(CT_ATAC     = "ATAC",     DynaTag_ATAC     = "ATAC",
              CT_H3K27ac  = "H3K27ac",  DynaTag_H3K27ac  = "H3K27ac",
              CT_H3K27me3 = "H3K27me3", DynaTag_H3K27me3 = "H3K27me3")

## nanoscope layout: results_<MARK>/PBMC_<LIBRARY>/fragments/fragments.tsv.gz
params_$frags <- vapply(names(lib_mark), function(n)
  file.path(params_$clusters_dir, paste0("results_", lib_mark[[n]]),
            paste0("PBMC_", n), "fragments", "fragments.tsv.gz"),
  character(1))
params_$frags <- as.list(params_$frags)
## ---- path preflight --------------------------------------------------------
## Fail here, loudly, with the offending path named -- not 500 lines later with
## an empty data.frame. `blacklist` is checked hardest: a missing blacklist does
## not error anywhere downstream, it just quietly stops removing the pile-up
## artifacts that are shared by every assay and therefore manufacture exactly
## the cross-mark concordance this vignette exists to explain.
.chk <- function(path, what, must_exist = TRUE, dir = FALSE) {
  ok <- if (dir) dir.exists(path) else file.exists(path)
  cat(sprintf("  %-12s %-6s %s\n", what, if (ok) "ok" else "MISSING", path))
  if (!ok && must_exist)
    stop("required path not found: ", path, " (", what, ")", call. = FALSE)
  ok
}
cat("path preflight\n")
.chk(params_$peaks_dir,  "peaks_dir",  dir = TRUE)
.chk(params_$encode_dir, "encode_dir", dir = TRUE)
.chk(params_$refs_dir,   "refs_dir",   dir = TRUE)
.chk(params_$blacklist,  "blacklist")
.chk(params_$helper,     "helper",     must_exist = FALSE)
for (n in names(params_$frags))
  .chk(params_$frags[[n]], "frag", must_exist = params_$run_fragment_analysis)

refs_dir <- params_$refs_dir
dir.create(params_$out_dir, recursive = TRUE, showWarnings = FALSE)

## ---- constants -------------------------------------------------------------
GENOME_BP <- 3088269832           # hg38 chr1-22,X,Y primary assembly
MAIN_CHR  <- paste0("chr", c(1:22, "X", "Y"))

peak_files <- c(
  CT_ATAC          = "PBMC_CT_ATAC_peaks.narrowPeak",
  DynaTag_ATAC     = "PBMC_DynaTag_ATAC_peaks.narrowPeak",
  CT_H3K27ac       = "PBMC_CT_H3K27ac_peaks.narrowPeak",
  DynaTag_H3K27ac  = "PBMC_DynaTag_H3K27ac_peaks.narrowPeak",
  CT_H3K27me3      = "PBMC_CT_H3K27me3_peaks.broadPeak",
  DynaTag_H3K27me3 = "PBMC_DynaTag_H3K27me3_peaks.broadPeak"
)
encode_files <- list(
  H3K27ac  = c("PBMC_H3K27ac_ENCSR105EMQ.bed", "PBMC_H3K27ac_ENCSR156XNC.bed",
               "PBMC_H3K27ac_ENCSR615HXA.bed"),
  H3K27me3 = c("PBMC_H3K27me3_ENCSR390SFH.bed", "PBMC_H3K27me3_ENCSR442BHP.bed",
               "PBMC_H3K27me3_ENCSR553XBX.bed")
)
meta <- data.frame(name = names(peak_files)) |>
  mutate(condition = sub("_.*", "", name), mark = sub("^[^_]+_", "", name))
cond_cols <- c(CT = "#1f78b4", DynaTag = "#e2711d")

## ---- namespace hygiene -----------------------------------------------------
## This is not optional. Two collisions silently break S4 dispatch on GRanges:
##   * purrr::reduce (attached by library(tidyverse)) masks the IRanges reduce()
##     generic. purrr then tries gr[[i]] and you get
##     "GRanges objects don't support [[, as.list(), lapply(), or unlist()".
##   * dplyr masks base::intersect/union/setdiff with generics whose .default
##     method is base's, which cannot handle S4 objects at all.
## Binding the generics directly makes the code independent of search-path order,
## so it behaves the same in a clean session and in one with Seurat + tidyverse
## already loaded.
gr_reduce    <- IRanges::reduce
gr_intersect <- BiocGenerics::intersect
gr_union     <- BiocGenerics::union
gr_setdiff   <- BiocGenerics::setdiff
gr_sort      <- BiocGenerics::sort

## report the masking so it is visible in the knitted output
for (fn in c("reduce", "intersect", "union", "setdiff", "sort")) {
  w <- utils::find(fn)
  if (length(w) > 1) message("NOTE: ", fn, "() is masked -- resolves to ", w[1],
                             "; using the explicit generic instead")
}
```

# 0. Preflight

Checks every path before anything expensive runs, because the fragment-level
sections take tens of minutes and failing at minute 40 on a typo is avoidable.

```{r preflight}
`%||%` <- function(a, b) if (is.null(a) || is.na(a)) b else a
chk <- function(label, path) data.frame(
  what = label, exists = file.exists(path %||% ""), path = path %||% "<unset>")

## peaks_dir auto-resolution. Depending on how the pseudobulk MACS2 run was laid
## out, the *_peaks.narrowPeak files sit either directly in IGV_clusters/ or in
## IGV_clusters/peaks/. Rather than make you guess and then debug an empty
## GRangesList, try the plausible candidates and take the one holding the files.
.cand <- unique(c(params_$peaks_dir,
                  file.path(params_$clusters_dir, "peaks"),
                  params_$clusters_dir))
.hits <- vapply(.cand, function(d) sum(file.exists(file.path(d, peak_files))),
                integer(1))
if (max(.hits) == 0L)
  stop("none of these directories contain the peak files:\n  ",
       paste(.cand, collapse = "\n  "), call. = FALSE)
if (.cand[which.max(.hits)] != params_$peaks_dir) {
  message("NOTE: peaks_dir repointed to ", .cand[which.max(.hits)],
          " (", max(.hits), "/6 peak files there vs ", .hits[[1]],
          "/6 at the configured path)")
  params_$peaks_dir <- .cand[which.max(.hits)]
}

pf <- rbind(
  chk("helper script", params_$helper),
  chk("blacklist",     params_$blacklist),
  do.call(rbind, lapply(names(peak_files), function(n)
    chk(paste("peaks:", n), file.path(params_$peaks_dir, peak_files[[n]])))),
  do.call(rbind, lapply(names(params_$frags), function(n)
    chk(paste("fragments:", n), params_$frags[[n]])))
)
if (any(!pf$exists)) {
  warning("missing inputs -- see the table below")
  print(pf[!pf$exists, ])
}
pf

## the fragment sections need BOTH the paths and the flag; report the gate state
message("run_fragment_analysis = ", params_$run_fragment_analysis,
        " | fragment paths set: ", sum(!is.na(unlist(params_$frags))), "/6")
```

# 1. Peak loading, cleaned

`reduce()` here is not cosmetic. It is what makes every later number well
defined.

```{r load-peaks}
bl <- if (file.exists(params_$blacklist)) {
  gr_reduce(import(params_$blacklist))
} else {
  warning("blacklist not found -- results will include artefact regions"); GRanges()
}

read_clean <- function(path) {
  stopifnot(file.exists(path))
  dt <- fread(path, header = FALSE, select = 1:3,
              col.names = c("chrom", "start", "end"))
  gr <- GRanges(dt$chrom, IRanges(dt$start + 1L, dt$end))     # BED is 0-based
  gr <- keepSeqlevels(gr, base::intersect(MAIN_CHR, seqlevels(gr)),
                      pruning.mode = "coarse")
  gr <- gr_reduce(gr)                                         # collapse summit dups
  if (length(bl)) gr <- gr[!overlapsAny(gr, bl)]              # drop, don't trim
  gr_sort(gr)
}

peaks <- lapply(peak_files, function(f) read_clean(file.path(params_$peaks_dir, f)))
enc   <- lapply(encode_files, function(fs)
  lapply(setNames(fs, sub("\\.bed$", "", fs)),
         function(f) read_clean(file.path(params_$encode_dir, f))))

## how much did cleaning remove? A large gap = the old numbers were inflated.
raw_n <- sapply(peak_files, function(f)
  nrow(fread(file.path(params_$peaks_dir, f), header = FALSE, select = 1L)))
data.frame(name = names(peaks), raw_intervals = raw_n,
           clean_intervals = sapply(peaks, length)) |>
  mutate(pct_removed = percent((raw_intervals - clean_intervals) / raw_intervals, 0.1))
```

On these data cleaning removes **19.7-21.0% of the narrow peaks** (ATAC, H3K27ac)
but only **0.7-0.8% of the broad ones**. That asymmetry matters for interpreting
the old numbers: `--call-summits` is only used for the narrow marks, so the
duplicate-interval bug inflated H3K27ac and ATAC counts by a fifth while leaving
H3K27me3 essentially untouched. Any previous narrow-vs-broad comparison was
therefore biased in a direction that had nothing to do with biology.

# 2. Peak statistics

Descriptive, and retained from the original — but read the **coverage** column
first. It is the denominator that every naive overlap statistic silently
rewards.

```{r peak-stats}
pstats <- data.frame(
  name     = names(peaks),
  n_peaks  = sapply(peaks, length),
  total_mb = sapply(peaks, function(g) sum(width(g)) / 1e6),
  mean_bp  = sapply(peaks, function(g) mean(width(g))),
  med_bp   = sapply(peaks, function(g) median(width(g)))
) |>
  left_join(meta, by = "name") |>
  mutate(coverage = total_mb * 1e6 / GENOME_BP)
pstats |> mutate(coverage = percent(coverage, 0.01))

ggplot(pstats, aes(mark, total_mb, fill = condition)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = cond_cols) +
  labs(y = "genome covered (Mb)", x = NULL,
       title = "Breadth, not peak number, drives naive overlap") +
  theme_classic(12)
```

# 3. Mark-exclusive ENCODE reference sets

This is the anchor. `ac_only` is the ENCODE H3K27ac union **minus** the
H3K27me3 union, and vice versa. Regions that are genuinely both are separated
into `bivalent` and excluded from the specificity statistic, so co-occupancy
cannot be mistaken for bleed.

Note the three H3K27ac ENCODE experiments disagree enormously among themselves
(4k/2.1 Mb vs 34k/26.3 Mb vs 30k/24.3 Mb). `ENCSR105EMQ` is an order of
magnitude smaller and is excluded from the union so it does not truncate it.

```{r encode-refsets}
enc_ac_union  <- gr_reduce(do.call(c, unname(enc$H3K27ac[
  base::setdiff(names(enc$H3K27ac), "PBMC_H3K27ac_ENCSR105EMQ")])))
enc_me3_union <- gr_reduce(do.call(c, unname(enc$H3K27me3)))

ac_only  <- gr_setdiff(enc_ac_union,  enc_me3_union)
me3_only <- gr_setdiff(enc_me3_union, enc_ac_union)

refsets  <- list(ac_only = ac_only, me3_only = me3_only)

## Export for the shell helper -- but NEVER overwrite files the driver wrote, so
## the fragment counts in sections 7-10 always refer to the same reference sets
## that produced them. If they already exist, compare sizes and warn on drift.
for (nm in names(refsets)) {
  f <- file.path(refs_dir, paste0(nm, ".bed"))
  if (!file.exists(f)) {
    export(refsets[[nm]], f, format = "BED")
  } else {
    on_disk <- sum(fread(f, select = 2:3, col.names = c("s", "e"))[, e - s])
    in_mem  <- sum(width(refsets[[nm]]))
    if (abs(on_disk - in_mem) / in_mem > 0.01)
      warning(nm, ".bed on disk covers ", round(on_disk/1e6, 1), " Mb but R built ",
              round(in_mem/1e6, 1), " Mb -- the cached fragment counts do not ",
              "correspond to these ranges. Delete refs/*.bed and re-run the driver.")
  }
}

data.frame(set = names(refsets),
           n_regions = sapply(refsets, length),
           mb = sapply(refsets, function(g) sum(width(g)) / 1e6),
           pct_genome = percent(sapply(refsets, function(g) sum(width(g))) / GENOME_BP, 0.01))
```

# 4. Specificity: the headline statistic

For each peak set and each reference set:

$$E = \frac{\text{fraction of the reference set covered by the peaks}}{\text{fraction of the genome covered by the peaks}}$$

Dividing by the peak set's own coverage is what makes this comparable across
libraries. $E = 1$ means "no better than scattering the same number of bases at
random", which is also exactly what a **uniform** background gives — so
$E > 1$ at `ac_only` for an H3K27me3 library cannot be ambient DNA or soup.
It requires a contaminant that is itself enriched in active chromatin.

The diagnostic is the **specificity ratio** $E[\text{on-target}] / E[\text{off-target}]$.

```{r specificity}
enrich_at <- function(pk, ref) {
  cov <- sum(width(pk)) / GENOME_BP
  obs <- sum(width(gr_intersect(ref, pk))) / sum(width(ref))
  c(coverage = cov, obs = obs, E = obs / cov)
}
spec <- do.call(rbind, lapply(names(peaks), function(n)
  do.call(rbind, lapply(names(refsets), function(r) {
    v <- enrich_at(peaks[[n]], refsets[[r]])
    data.frame(name = n, set = r, coverage = v[["coverage"]],
               obs_frac = v[["obs"]], E = v[["E"]])
  })))) |> left_join(meta, by = "name")

spec_tab <- spec |>
  select(name, mark, condition, set, E) |>
  pivot_wider(names_from = set, values_from = E, names_prefix = "E_") |>
  mutate(
    on_target  = ifelse(mark == "H3K27me3", E_me3_only, E_ac_only),
    off_target = ifelse(mark == "H3K27me3", E_ac_only,  E_me3_only),
    specificity_ratio = on_target / off_target
  )
spec_tab |> mutate(across(where(is.numeric), ~round(.x, 2)))

ggplot(filter(spec, set != "bivalent"), aes(set, E, fill = condition)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 1, linetype = 2) +
  facet_wrap(~mark, scales = "free_y") +
  scale_fill_manual(values = cond_cols) +
  labs(y = "enrichment over own genome coverage", x = NULL,
       title = "Dashed line = random expectation = uniform background") +
  theme_bw(12)
```

**How to read this on the PBMC data.** H3K27ac is clean in both protocols
($E[ac\_only] \approx 16$, and near-random at `me3_only`). The entire problem
localises to **DynaTag H3K27me3**: off-target $E[ac\_only]$ rises from 0.80 (CT,
i.e. *depleted*, as a repressive mark should be) to 2.34, and the specificity
ratio collapses from 7.96 to 1.71.

The on-target value also appears to fall (6.36 → 4.01), but **do not interpret
that as lost sensitivity** — §4b shows it is entirely an artifact of comparing
sets with 11% vs 21% genome coverage. At matched coverage the on-target
enrichment is identical (6.36 vs 6.27). The correct model is therefore *purely
additive* off-target signal on top of intact on-target signal, not dilution of
it.

The `bivalent` column is a useful sanity check rather than a discriminator:
CT H3K27me3 scores 8.20 there, *above* its own `me3_only` value of 6.36, which is
what real bivalent promoters should do and confirms the reference sets are
behaving. DynaTag H3K27me3 scores 4.58, barely different from its 4.01 at
`me3_only` — consistent with a library that has lost positional information
rather than gained bivalent sensitivity.

A crude mixing estimate, treating the library as a mixture of a specific
component and an active-chromatin-targeted contaminant,
$\alpha = (E_{obs} - E_{spec}) / (E_{contam} - E_{spec})$, puts the contaminant
near 5–10%. Treat that as order-of-magnitude only: peak coverage saturates and
is threshold-dependent, so §7 recomputes it on fragments, which is the version
to trust.

# 4b. Threshold sweep: does filtering more fix it?

The obvious next move is to tighten the MACS q-value cutoff — the calls above use
$q<0.01$ narrow / $q<0.05$ broad, and 21% of the genome called as H3K27me3 looks
like an over-permissive threshold. This section tests that, and the answer is
**no: filtering makes the specificity worse, not better.** That is worth knowing
before spending time on it.

Two things are computed. First a sweep of $-\log_{10}q$, recording specificity at
each cutoff. Second — and this is the one that settles it — a **coverage-matched**
comparison, thresholding the broader library until it covers the same number of
base pairs as the narrower one, which removes breadth as a variable entirely.

```{r threshold-sweep}
## needs the q-value column, so re-read unmerged and threshold before reducing
peaks_at <- function(path, qthr) {
  dt <- fread(path, header = FALSE)
  dt <- dt[dt$V9 >= qthr & dt$V1 %in% MAIN_CHR]
  if (!nrow(dt)) return(GRanges())
  gr <- gr_reduce(GRanges(dt$V1, IRanges(dt$V2 + 1L, dt$V3)))
  if (length(bl)) gr <- gr[!overlapsAny(gr, bl)]
  gr
}
spec_of <- function(gr, mark) {
  if (!length(gr)) return(c(mb = 0, cov = 0, E_ac = NA, E_me3 = NA, ratio = NA))
  cov  <- sum(width(gr)) / GENOME_BP
  E_ac <- (sum(width(gr_intersect(ac_only,  gr))) / sum(width(ac_only)))  / cov
  E_me <- (sum(width(gr_intersect(me3_only, gr))) / sum(width(me3_only))) / cov
  on   <- if (mark == "H3K27me3") E_me else E_ac
  off  <- if (mark == "H3K27me3") E_ac else E_me
  c(mb = sum(width(gr)) / 1e6, cov = cov, E_ac = E_ac, E_me3 = E_me, ratio = on / off)
}

qgrid <- c(1.3, 2, 3, 5, 10, 20, 50)
sweep <- do.call(rbind, lapply(names(peak_files), function(n) {
  m <- meta$mark[meta$name == n]; if (m == "ATAC") return(NULL)
  do.call(rbind, lapply(qgrid, function(q) {
    v <- spec_of(peaks_at(file.path(params_$peaks_dir, peak_files[[n]]), q), m)
    data.frame(name = n, mark = m, condition = meta$condition[meta$name == n],
               qthr = q, t(v))
  }))
}))
sweep |> mutate(across(c(mb, cov, E_ac, E_me3, ratio), ~round(.x, 3)))

ggplot(sweep, aes(qthr, ratio, colour = condition)) +
  geom_line() + geom_point() +
  scale_x_log10() + scale_y_log10() +
  geom_hline(yintercept = 1, linetype = 2) +
  facet_wrap(~mark, scales = "free_y") +
  scale_colour_manual(values = cond_cols) +
  labs(x = expression(-log[10]~q~threshold), y = "specificity ratio (on / off target)",
       title = "A specific assay gets MORE specific as you filter") +
  theme_bw(12)
```

```{r coverage-match}
## Bisect the threshold until the broader library matches the narrower one's bp.
match_coverage <- function(path, mark, target_bp, lo = 1.3, hi = 100, iter = 20) {
  for (i in seq_len(iter)) {
    mid <- (lo + hi) / 2
    b <- sum(width(peaks_at(path, mid)))
    if (b > target_bp) lo <- mid else hi <- mid
  }
  list(qthr = mid, gr = peaks_at(path, mid))
}
target <- sum(width(peaks[["CT_H3K27me3"]]))
mm <- match_coverage(file.path(params_$peaks_dir, peak_files[["DynaTag_H3K27me3"]]),
                     "H3K27me3", target)
rbind(
  data.frame(library = "CT_H3K27me3 (as called)", qthr = 1.3,
             t(spec_of(peaks[["CT_H3K27me3"]], "H3K27me3"))),
  data.frame(library = "DynaTag_H3K27me3 (coverage-matched)", qthr = mm$qthr,
             t(spec_of(mm$gr, "H3K27me3")))
) |> mutate(across(where(is.numeric), ~round(.x, 3)))
```

**Result on the PBMC data, and it is unambiguous.**

*CT H3K27me3 behaves as a specific assay must.* Tightening the cutoff raises the
specificity ratio monotonically — 7.96 → 8.88 → 12.4 → 23.4 → 64.6 → 191 across
the grid — because $E[ac\_only]$ *falls* (0.80 → 0.13) while $E[me3\_only]$ rises
(6.36 → 24.6). The strongest peaks are the most on-target ones.

*DynaTag H3K27me3 does the opposite.* The ratio is flat and then **inverts**:
1.71 → 1.71 → 1.67 → 1.48 → 1.43 → 0.77 → 0.07. $E[ac\_only]$ climbs steeply with
the threshold, 2.34 → 33.3. At $q$-thresholds above ~20 the *most statistically
significant* "H3K27me3" peaks in this library sit preferentially at
H3K27ac-exclusive regions — and $E[ac\_only] = 33$ is essentially the value the
**ATAC** libraries score (32.3 and 33.3).

So the off-target component is not weak background scraping the cutoff. It carries
the **highest pileup in the library**, which is why filtering concentrates it.
Raising the threshold selects *for* the artifact.

The coverage-matched row is the cleanest statement of the whole problem. At 351 Mb
vs 352 Mb — matched to within 0.3% — CT and DynaTag recover H3K27me3 domains
*equally well* ($E[me3\_only]$ 6.36 vs 6.27), while off-target differs five-fold
($E[ac\_only]$ 0.80 vs 3.97). DynaTag H3K27me3 is intact on-target signal **plus**
an additive, high-pileup, accessibility-tracking component. Nothing has been lost;
something has been added.

*Control.* The same sweep on H3K27ac behaves identically in both protocols
(specificity improving monotonically, $E[ac\_only]$ reaching 63 and 69 at
$q>100$), so neither the sweep machinery nor the DynaTag protocol as a whole is
broken — the defect is specific to the H3K27me3 channel.

**Practical answer to "should I filter more?"** No. Not for this purpose. Peak
breadth is a symptom, not the disease, and thresholding trades recall away while
*enriching* the contaminant. The high-pileup, ATAC-like signature points the
mechanism at untethered pA-Tn5 or barcode hopping **from the ATAC channel**, both
of which are fixed upstream — at the bench or at demultiplexing — not by a cutoff.
§8 and §10 discriminate between them.

# 5. Corrected bp-level Jaccard

$J = |A \cap B|_{bp} / |A \cup B|_{bp}$, bounded in $[0,1]$, on reduced ranges.

```{r jaccard}
jac <- function(a, b) {
  i <- sum(width(gr_intersect(a, b))); u <- sum(width(gr_union(a, b)))
  if (u == 0) 0 else i / u
}
enc_flat <- unlist(enc, recursive = FALSE)
names(enc_flat) <- sub("^.*_(ENCSR[0-9A-Z]+)$", "ENC_\\1", names(enc_flat))
all_sets <- c(peaks, enc_flat)
J <- outer(seq_along(all_sets), seq_along(all_sets),
           Vectorize(function(i, j) jac(all_sets[[i]], all_sets[[j]])))
dimnames(J) <- list(names(all_sets), names(all_sets))

as.data.frame(as.table(J)) |>
  setNames(c("a", "b", "J")) |>
  ggplot(aes(a, b, fill = J)) + geom_tile() +
  geom_text(aes(label = sprintf("%.2f", J)), size = 2.4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(x = NULL, y = NULL, title = "bp-level Jaccard (reduced, blacklist-filtered)") +
  theme_minimal(10) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 6. Recall and precision, against the achievable ceiling

Jaccard is the wrong single number when two sets differ several-fold in
breadth: it is penalised by the size mismatch itself. Recall
($|A \cap B| / |B|$, how much of ENCODE is recovered) and precision
($|A \cap B| / |A|$) separate those.

Critically, the ceiling is not 1 — it is **ENCODE against ENCODE**. On these
data the H3K27me3 replicates agree with each other at $J = 0.083–0.219$, and
the observed CT/DynaTag-vs-ENCODE values (0.086–0.132) sit *inside* that
range. There is no concordance deficit to explain; the reference is simply not
self-consistent enough to define one.

```{r recall-precision}
rp <- function(a, b) {
  i <- sum(width(gr_intersect(a, b)))
  c(recall = i / sum(width(b)), precision = i / sum(width(a)), J = jac(a, b))
}
rp_tab <- do.call(rbind, lapply(names(peaks), function(n) {
  m <- meta$mark[meta$name == n]
  if (!m %in% names(enc)) return(NULL)
  do.call(rbind, lapply(names(enc[[m]]), function(e) {
    v <- rp(peaks[[n]], enc[[m]][[e]])
    data.frame(name = n, encode = e, recall = v[["recall"]],
               precision = v[["precision"]], J = v[["J"]])
  }))
}))

## the ceiling: every ENCODE-vs-ENCODE pair within a mark
ceiling_tab <- do.call(rbind, lapply(names(enc), function(m) {
  s <- enc[[m]]; if (length(s) < 2) return(NULL)
  cb <- combn(names(s), 2)
  data.frame(mark = m, pair = apply(cb, 2, paste, collapse = " vs "),
             J = apply(cb, 2, function(p) jac(s[[p[1]]], s[[p[2]]])))
}))
list(observed = rp_tab |> mutate(across(where(is.numeric), ~round(.x, 3))),
     encode_ceiling = ceiling_tab |> mutate(J = round(J, 3)))
```

More aggressive peak filtering trades recall for precision at roughly 1:1 and
does **not** rescue Jaccard — on H3K27ac, keeping the top 20k peaks by
q-value moves $J$ from 0.120 to 0.241 while recall drops 65%→45%. Filtering is
therefore not the fix; it changes which error you make.

```{r jaccard-vs-topN, eval=FALSE}
## Set eval=TRUE to reproduce. Re-reads the unmerged file for the q-value column.
sweep_topn <- function(f, ns = c(2e4, 5e4, 1e5, Inf)) {
  dt <- fread(f, header = FALSE)
  setorder(dt, -V9)                                   # V9 = -log10(q)
  lapply(ns, function(n) {
    d <- dt[seq_len(min(n, .N))]
    gr <- gr_reduce(GRanges(d$V1, IRanges(d$V2 + 1L, d$V3)))
    gr <- keepSeqlevels(gr, base::intersect(MAIN_CHR, seqlevels(gr)),
                        pruning.mode = "coarse")
    if (length(bl)) gr <- gr[!overlapsAny(gr, bl)]
    gr
  }) |> setNames(paste0("top", ns))
}
```

# 7. Fragment-level signal budget

Peak coverage cannot answer "where is the extra signal going", because once a
region is called a peak, additional fragments there change nothing. Fragments
can. This is the section that turns "DynaTag H3K27me3 has 135M fragments vs
CT's 52M" into a statement about *where* those 83M extra fragments land.

The within-library **share** matters more than the absolute count, and is
depth-independent by construction: in nanoCT both marks share one tube and one
sequencing run, split only by antibody barcode. H3K27me3 going from 36.1% to
54.6% of the library — a 2.13× shift in the me3:ac ratio — cannot be produced
by sequencing depth, which scales both marks together. That confirms the
protocol, not the sequencer, changed what was captured.

```{r fragment-budget}
run_helper <- function(args) {
  if (!file.exists(params_$helper))
    stop("helper script not found at '", params_$helper, "'.\n",
         "  It must exist on the machine where R is running. If you wrote it in\n",
         "  Codon, scp it to the cluster and set params_$helper to an ABSOLUTE path\n",
         "  -- a relative path resolves against the .Rmd's directory when knitting,\n",
         "  which is not where you think it is.", call. = FALSE)
  ## capture BOTH streams and check the exit status. Without this a failing
  ## bedtools call returns character(0) and the chunk silently does nothing.
  out <- suppressWarnings(
    system2("bash", c(shQuote(params_$helper), args), stdout = TRUE, stderr = TRUE))
  st <- attr(out, "status")
  if (!is.null(st) && st != 0)
    stop("helper failed (exit ", st, ") on: ", paste(args, collapse = " "),
         "\n", paste(out, collapse = "\n"), call. = FALSE)
  out
}
if (params_$run_fragment_analysis) {
  ## `profile` emits budget_*.tsv AND widths_*.tsv from a single sort of the
  ## fragments file. Calling `budget` and `widths` separately sorts it five times.
  for (n in names(params_$frags)) {
    f <- params_$frags[[n]]; if (is.na(f)) next
    message("profiling ", n, " ...")
    run_helper(c("profile", "-o", refs_dir, "--name", n, "--frag", f,
                 if (isTRUE(params_$frags_presorted)) "--presorted"))
  }
}
budget_files <- list.files(refs_dir, "^budget_.*\\.tsv$", full.names = TRUE)
if (length(budget_files)) {
  budget <- rbindlist(lapply(budget_files, fread)) |>
    left_join(meta, by = "name")
  tot <- budget |> filter(partition == "total") |> select(name, total = n_fragments)
  bud <- budget |> filter(partition != "total") |> left_join(tot, by = "name") |>
    mutate(frac = n_fragments / total) |> filter(partition != "bivalent")
  print(bud |> mutate(frac = percent(frac, 0.01)))

  ggplot(bud, aes(partition, frac, fill = condition)) +
    geom_col(position = "dodge") +
    facet_wrap(~mark) + scale_y_continuous(labels = percent) +
    scale_fill_manual(values = cond_cols) +
    labs(y = "fraction of library fragments", x = NULL,
         title = "Signal budget: where the fragments actually land") +
    theme_bw(12)
} else {
  cat("No budget_*.tsv in", refs_dir, "\n")
  cat("  fragment paths set:      ", sum(!is.na(unlist(params_$frags))), "/6\n")
  cat("  run_fragment_analysis:   ", params_$run_fragment_analysis, "\n")
  ## Show what IS there. An empty listing means refs_dir is the wrong directory
  ## (the usual cause); a listing with refsets_summary.tsv but no budget_*.tsv
  ## means the driver died partway through its section 2.
  have <- list.files(refs_dir)
  cat("  refs_dir contains ", length(have), " file(s)",
      if (length(have)) ":" else " -- wrong directory?", "\n", sep = "")
  if (length(have)) cat(paste0("    ", head(have, 25), collapse = "\n"), "\n")
  cat("Both are required. Filling params_$frags alone does not trigger the run.\n")
}
```

The excess is not one thing. Two observations force at least two components:
the contaminant fraction implied by `ac_only` enrichment (~5–10%, i.e. 7–14M
fragments) is far too small to account for 83M extra fragments, and peak
breadth grows 352 → 658 Mb while `me3_only` enrichment falls more steeply than
a 10% dilution predicts. So expect a small active-chromatin-targeted component
**plus** a larger broadly-distributed one.

# 8. Fragment widths per partition — the mechanistic test

This is what distinguishes the candidate mechanisms, because they leave
different length signatures:

- **untethered pA-Tn5** tagmenting open chromatin → sub-100 bp,
  ATAC-like, concentrated in `ac_only`;
- **barcode hopping or cross-reactivity** → a normal nucleosomal ladder
  (~150 / 300 / 450 bp) in the wrong place, indistinguishable by length from
  real signal (so §9 is needed for those);
- **over-digestion / broad diffuse capture** → ladder degraded everywhere,
  including on-target.

```{r fragment-widths}
## widths_*.tsv were already written by the `profile` call in section 7.
width_files <- list.files(refs_dir, "^widths_.*\\.tsv$", full.names = TRUE)
if (length(width_files)) {
  wdt <- rbindlist(lapply(width_files, fread)) |> left_join(meta, by = "name")
  wdt <- wdt |> group_by(name, partition) |> mutate(dens = count / sum(count)) |> ungroup()

  print(ggplot(filter(wdt, width <= 800, partition != "all" & partition != "bivalent"),
               
               aes(width, dens, colour = condition)) +
    geom_line() +
    geom_vline(xintercept = c(150, 300, 450), linetype = 3, colour = "grey40") +
    facet_grid(mark ~ partition) +
    scale_colour_manual(values = cond_cols) +
    labs(x = "fragment length (bp)", y = "density",
         title = "Dotted lines = mono / di / tri-nucleosome") +
    theme_bw(11))

  ## one number per library: sub-nucleosomal share, on- vs off-target
  wdt |> filter(partition != "all") |>
    group_by(name, mark, condition, partition) |>
    summarise(sub_100bp = sum(count[width < 100]) / sum(count),
              nucleosomal_150_250 = sum(count[width >= 150 & width <= 250]) / sum(count),
              .groups = "drop") |>
    mutate(across(where(is.numeric), ~percent(.x, 0.1)))
} else cat("No widths_*.tsv found.\n")
```

# 9. Depth-matched recomputation

The only fully convincing control for a depth difference: subsample the deeper
library to the shallower library's exact fragment count and repeat §4. Uniform
Bernoulli thinning preserves the spatial distribution, so anything that
survives is not a depth artefact.

```{r depth-match, eval=FALSE}
## eval=FALSE because it needs a re-run of MACS2 on the subsampled fragments.
## 51,956,130 = CT H3K27me3 total.
ds <- file.path(refs_dir, "DynaTag_H3K27me3.ds.tsv.gz")
params_$bin_dir <- "/data/bin/bedtools2/bin"
run_helper <- function(args) {
  if (!file.exists(params_$helper))
    stop("helper script not found at '", params_$helper, "'", call. = FALSE)
  env <- character(0)
  if (nzchar(params_$bin_dir))
    env <- paste0("PATH=", params_$bin_dir, ":", Sys.getenv("PATH"))
  out <- suppressWarnings(
    system2("bash", c(shQuote(params_$helper), args),
            stdout = TRUE, stderr = TRUE, env = env))
  st <- attr(out, "status")
  if (!is.null(st) && st != 0)
    stop("helper failed (exit ", st, ") on: ", paste(args, collapse = " "),
         "\n", paste(out, collapse = "\n"), call. = FALSE)
  out
}
## then re-call peaks with the same broad flags used originally:
##   macs2 callpeak -f BEDPE --keep-dup all --broad --broad-cutoff 0.05 \
##     --max-gap 1000 --llocal 100000 -q 0.01 -g hs -t <bedpe from ds>
## and add the result to `peaks` before re-running the specificity chunk.
run_helper(c("downsample", "--frag", params_$frags[["DynaTag_H3K27me3"]],
             "-N", "51956130", "--out", ds))
file.exists(ds)
```

# 10. Barcode-hopping test

A single tagmentation event carries exactly one antibody barcode. So a fragment
with **identical** `(chrom, start, end)` *and* the same cell barcode in two
modality files is a mis-assignment, not biology. The null permutes the cell
barcode column — destroying the cell pairing while preserving every marginal
distribution — so `obs/exp` isolates the pairing itself. Restricted to one
chromosome for tractability; the rate is a genome-wide property.

An `obs/exp` near 1 exonerates hopping and points at cross-reactivity or
untethered Tn5 instead. Substantially above 1 implicates index swapping, which
is fixable at the sequencing/demultiplexing step rather than at the bench.

```{r collision, eval=FALSE}

run_helper(c("collision", "-o", refs_dir, "--name", "CT",
             "--frag-a", params_$frags$CT_H3K27ac,
             "--frag-b", params_$frags$CT_H3K27me3,
             "--chrom", "chr1", "--perms", "5"))
run_helper(c("collision", "-o", refs_dir, "--name", "DynaTag",
             "--frag-a", params_$frags$DynaTag_H3K27ac,
             "--frag-b", params_$frags$DynaTag_H3K27me3,
             "--chrom", "chr1", "--perms", "5"))
## §4b implicates the ATAC channel specifically (the off-target signature matches
## the ATAC libraries' enrichment), so test that pair in both conditions too.
for (cond in c("CT", "DynaTag"))
  run_helper(c("collision", "-o", refs_dir, "--name", paste0(cond, "_ATACvME3"),
               "--frag-a", params_$frags[[paste0(cond, "_ATAC")]],
               "--frag-b", params_$frags[[paste0(cond, "_H3K27me3")]],
               "--chrom", "chr1", "--perms", "5"))
```

```{r collision-report}
cf <- list.files(refs_dir, "^collision_.*\\.tsv$", full.names = TRUE)
if (length(cf)) rbindlist(lapply(cf, fread)) |> pivot_wider(names_from = metric,
                                                            values_from = value) else
  cat("No collision_*.tsv found.\n")
```

# 11. Fine-scale correlation inside the peak union

The 5-kb genome-wide correlation you already have is dominated by A/B
compartment structure, which **any** two chromatin tracks share — that is why
even the ENCODE H3K27ac/H3K27me3 pair only floors at 0.13–0.31 rather than
going negative, and why your 0.48 (CT) / 0.59 (DynaTag) are not directly
interpretable as cross-reactivity.

At 500 bp inside the peak union, genuine H3K27ac and H3K27me3 should be
**anti**-correlated, because the marks are locally mutually exclusive. A
positive fine-scale Spearman is much harder to explain away.

Two caveats on the existing heatmap: `CT_ATAC` vs `DynaTag_ATAC` = 0.99 with
off-diagonal rows identical to two decimals is suspicious — the peak Jaccard is
0.740, so the files are not identical, but md5 the bigwigs. And deepTools RPKM
does not remove a differing *background fraction*, so quantile-normalise or use
fold-over-genome-background before comparing across protocols.

```{r fine-scale, eval=FALSE}
params_$multibigwigsummary <- "/home/mattia/miniconda3_n/envs/nanoscope_general/bin/multiBigwigSummary"
## --- 0. deepTools must be reachable from R -----------------------------
## R/RStudio Server gets a minimal non-login PATH. Prepend the conda bin
## that holds multiBigwigSummary ONCE, then Sys.which() also becomes
## authoritative for the rest of the session.
if (nzchar(params_$bin_dir %||% "") &&
    !grepl(params_$bin_dir, Sys.getenv("PATH"), fixed = TRUE))
  Sys.setenv(PATH = paste(params_$bin_dir, Sys.getenv("PATH"), sep = ":"))
mbs <- Sys.which("multiBigwigSummary")
use_dt <- nzchar(mbs)
if (!use_dt)
  message("multiBigwigSummary not on PATH -- using the rtracklayer fallback.\n",
          "  (To use deepTools instead, put its bin in params_$bin_dir;\n",
          "   colon-separate several dirs. The fallback is equivalent.)")

## --- 1. 500 bp tiles over the peak union ------------------------------
## multiBigwigSummary's BED-file mode gives ONE value per BED interval and
## does NOT accept --binSize (that flag belongs to `bins` mode only, and
## passing it here is an argparse error). To get 500 bp resolution we tile
## the union ourselves and hand deepTools the tiles.
peak_union <- gr_reduce(do.call(c, unname(peaks[c("CT_H3K27ac", "CT_H3K27me3",
                                                  "DynaTag_H3K27ac", "DynaTag_H3K27me3")])))
tiles <- unlist(GenomicRanges::tile(peak_union, width = 500L))
tiles <- tiles[BiocGenerics::width(tiles) >= 250L]   # drop ragged remainders
mcols(tiles) <- NULL
union_bed <- file.path(refs_dir, "peak_union_500bp.bed")
rtracklayer::export(tiles, union_bed, format = "BED")
message(sprintf("peak union: %d intervals, %.1f Mb -> %d tiles",
                length(peak_union), sum(BiocGenerics::width(peak_union))/1e6,
                length(tiles)))

## --- 2. locate the bigwigs --------------------------------------------
bw_all <- list.files(params_$clusters_dir, pattern = "\\.(bw|bigwig|bigWig)$",
                     recursive = TRUE, full.names = TRUE)
bw <- vapply(names(lib_mark), function(n) {
  hit <- bw_all[grepl(paste0("(^|[^A-Za-z0-9])", n, "([^A-Za-z0-9]|$)"),
                      basename(bw_all))]
  if (length(hit) != 1L) NA_character_ else hit
}, character(1))
if (anyNA(bw)) {
  cat("bigwigs found under clusters_dir:\n"); print(basename(bw_all))
  stop("could not uniquely match a bigwig for: ",
       paste(names(bw)[is.na(bw)], collapse = ", "),
       "\nSet `bw` by hand as a named character vector and re-run.")
}
print(data.frame(library = names(bw), file = basename(bw), row.names = NULL))

## --- 3. quantify the tiles -------------------------------------------
if (use_dt) {
  npz <- file.path(refs_dir, "fine.npz")
  tab <- file.path(refs_dir, "fine.tab")
  args <- c("BED-file",
            "-b", unname(bw),
            "--labels", names(bw),
            "--BED", union_bed,
            "-p", "8",
            "-o", npz,
            "--outRawCounts", tab)
  st <- system2(mbs, shQuote(args), stdout = TRUE, stderr = TRUE)
  cat(st, sep = "\n")
  if (!file.exists(tab))
    stop("multiBigwigSummary did not write ", tab, " -- see the output above.")
  ## --outRawCounts writes a commented, quoted header: #'chr' 'start' 'end' ...
  ## and its row order is NOT guaranteed to match the input BED, so re-key.
  fine <- fread(tab)
  setnames(fine, gsub("^#|'", "", names(fine)))
  mat <- as.matrix(fine[, -(1:3)])
} else {
  ## rtracklayer fallback: identical quantity (mean bigwig signal per tile),
  ## computed in-process. size = 1L => one summary value per interval.
  ## defaultValue = 0 makes uncovered tiles 0 rather than NaN.
  mat <- vapply(bw, function(f) {
    sm <- rtracklayer::summary(rtracklayer::BigWigFile(f),
                              which = tiles, size = 1L,
                              type = "mean", defaultValue = 0)
    as.numeric(unlist(lapply(sm, function(g) score(g)[1L])))
  }, numeric(length(tiles)))
  colnames(mat) <- names(bw)
  fine <- data.table(chr   = as.character(seqnames(tiles)),
                     start = start(tiles) - 1L,
                     end   = end(tiles))
}
mat[!is.finite(mat)] <- 0                      # deepTools emits nan over gaps
keep <- rowSums(mat) > 0                       # all-zero tiles carry no signal
message(sprintf("%d/%d tiles retained (non-zero)", sum(keep), nrow(mat)))
round(cor(log1p(mat[keep, , drop = FALSE]), method = "spearman"), 3)
```

# 12. Genomic annotation

Retained, and now on cleaned ranges — the previous version annotated duplicated
summit intervals, which double-counted the promoter fraction.

```{r annotate, eval=FALSE}
suppressPackageStartupMessages({
  library(ChIPseeker); library(TxDb.Hsapiens.UCSC.hg38.knownGene)
})
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
ann  <- lapply(peaks, annotatePeak, TxDb = txdb,
               tssRegion = c(-2000, 2000), verbose = FALSE)
plotAnnoBar(ann)
```

# Statistical caveats

- **n = 1 pseudobulk per condition.** There is no replicate variance, so no
  valid differential test exists. Report fold-change rankings, never p-values.
  If replicates are unavailable, split cells into pseudo-replicates and treat
  the result as a stability check, not inference.
- **Post-dedup fragment counts assume comparable saturation.** Compare
  `preseq lc_extrap` curves and duplication rates for the four libraries before
  interpreting the 52M vs 135M difference as captured complexity.
- **MACS2 `--broad` is a poor caller for very deep broad-mark pseudobulks** —
  it is what produced 661 Mb of H3K27me3 "peaks" (21% of the genome). SEACR or
  a binned-coverage caller with an explicit background model is more
  appropriate, and the §4 statistic should be recomputed on those calls.
- The mixture estimate in §4 is peak-coverage-based and therefore
  threshold-dependent; §7 is the fragment-level version.

```{r session}
sessionInfo()
```
