---
title: "nanoCTAR RNA all together"
author: "Mattia Zaghi"
date: "2025-08-28"
output: html_document
---

## 1. Data loading (metadata and fragments)
In this section we are going to load the input files for each modality and sample, create a fragment object for each experiment, generate a bins x cell matrix for each experiment and finally generate one Seurat object for experiment.

* Inputs of this section are the ```fragments.tsv.gz``` and ```metadata.csv``` files.\
* Output of his section is a Seurat object for each experiment (6 in this case [2 samples * 3 modalities]).\

First, load all the required libraries.

```{r setup2}
rm(list = ls())
library(Signac)
library(Seurat)
library(GenomicRanges)
library(future)
library(stringr)
library(ggplot2)
library(gghalves)
library(ggpubr)
library(EnsDb.Mmusculus.v79)
library(BSgenome.Mmusculus.UCSC.mm10)
library(ComplexUpset)
library(regioneR)
library(scales)
library(ggVennDiagram)
library(hdf5r)
```



### 1.1 Setting some initial variables
First, define some variables that will be used across the entire vignette. Modify them according to your data.\

**Extremely important:**

1. set as ```repodir``` the directory where the [nanoscope Github repo](https://github.com/bartosovic-lab/nanoscope/blob/main/README.md#clone-github-repository) is. This is required in order to upload all the custom functions needed to run this vignette\

```{r init_var1 }
# directory where the nanoscope repo was cloned
repodir <- "/home/mattia/nanoCTAR/" # change accordingly
source(paste0(repodir,"scripts/functions_scCT.R"))
```


2. set as ```wd``` the output root directory.\
* If you are following [our tutorial](https://github.com/bartosovic-lab/nanoscope), please place the nanoscope output directories ```sample_P23209``` and ```sample_P24004``` in a directory called ```result``` and set it as working directory ```wd <- "~/NatProt/nanoscope/results/"```\
* If you have skipped the tutorial and directly downloaded the toy data from [Zenodo](https://zenodo.org/record/7759368#.ZBrMcC-B3Eo), your ```wd``` will be ```/path/to/nanoscope_toy_data/results/```\
* If you are running this vignette on your personal data please ensure that your ```wd``` is formatted as follow:\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;- ```fragments.tsv.gz``` is expected to be in a folder with this path: ```wd/{sample_name}/{modality_barcode}/cellranger/outs/```\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;- ```metadata.csv``` is expected to be in a folder with this path: ```wd/{sample_name}/{modality_barcode}/cell_picking/```\
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;-Note that this vignette assumes the same modalities to have the same barcodes across the different samples\

**Please note that nothing but the directories containing the input data for each samples must be stored in ```wd```**
```{r init_var2 }
# directory where the nanoscope output files are stored
require("knitr")
opts_knit$set(root.dir = "/proj/user/mattia/Embryo_2/")# change accordingly

#wd <- "~/NatProt/nanoscope/results" # change accordingly
#setwd(wd)


```


3. set other variables (genome version, assay type, etc)
```{r init_var3 }
# Set the used genome version
genome <- "mm10"
# Genome annotations
genome_ann <- EnsDb.Mmusculus.v79
# BSgenome
bs_genome <- BSgenome.Mmusculus.UCSC.mm10
# Assay (peaks or bins)
assay <- "bins"
# Samples
#samples <- dir()
samples<-c("nanoCTAR_e13_1","nanoCTAR_e13_2","nanoCTAR_e13_K4me3_1","nanoCTAR_e13_K4me3_2","nanoCTAR_e15_1","nanoCTAR_e15_2",
           "e15_nanoCTR_1","e15_nanoCTR_2")
# Modalities
#modalities <- dir(samples[1]) # this assumes sample1, sample2, sampleN share the same modalities and barcodes!
modalities<-c("RNA_AAAAGGGG")
# Minimum number of cells a feature must be detected in to be included in the Seurat obj
min_cell <- 1
# Minium number of features a cell must have to be included in the Seurat obj
min_feat <- 1
```



###1.3 Create one Counts RNA object for each experiment
```{r}
##### RNA
# Create one RNA object per each experiment iterating across the sample list. To do so you need change the working direcotry inside this chunk.
# Samples
samples <- samples
counts_RNA.ls <- list()

for (smpl in samples) {
   for (mod in modalities) {
  cat("Loading counts for sample",smpl,"\n")

    # Load fragments
counts_RNA.ls[[paste0(mod,"_",smpl)]]<-  Read10X_h5(paste0("/proj/user/mattia/Embryo_2/",smpl,"/RNA_AAAAGGGG/cellranger/outs/raw_feature_bc_matrix.h5"))

   }
}


```

Now we will create an RNA object list list and perform some QC before unifying all the objects
```{r create_obj}
obj.ls_RNA<-list()
for (experim in names(counts_RNA.ls)) {
  # Get sample and modality name from the experiment variable
  smpl <- str_split_fixed(experim,"_",3)[,3]
  modality <- str_split_fixed(experim,"_",2)[,1]

  
  # Create the object
  obj.ls_RNA[[paste0(experim)]]<- CreateSeuratObject(counts = counts_RNA.ls[[paste0(experim)]],
                                            assay = "RNA",
                                            project = smpl,  min.cells = 1,
  min.features = 1)

    obj.ls_RNA[[paste0(experim)]]@meta.data$barcode_RNA<- colnames(obj.ls_RNA[[paste0(experim)]])
    obj.ls_RNA[[paste0(experim)]][["percent.mt"]] <- PercentageFeatureSet(obj.ls_RNA[[paste0(experim)]], pattern = "^mt-")
    # Add some metadata of origin
  obj.ls_RNA[[paste0(experim)]]$dataset <- experim
  obj.ls_RNA[[paste0(experim)]]$modality <- modality
  obj.ls_RNA[[paste0(experim)]]$sample <- smpl
}

```

UNique RNA merged object


```{r}
# Merge list
combined.obj_RNA <- list()

    for (mod in modalities) {

  # merge objects among the different samples
  combined.obj_RNA[[mod]] <- merge(obj.ls_RNA[[paste0(mod,"_",samples[1])]],
                                  y = c(obj.ls_RNA[[paste0(mod,"_",samples[2])]],obj.ls_RNA[[paste0(mod,"_",samples[3])]],obj.ls_RNA[[paste0(mod,"_",samples[4])]],obj.ls_RNA[[paste0(mod,"_",samples[5])]],obj.ls_RNA[[paste0(mod,"_",samples[6])]],obj.ls_RNA[[paste0(mod,"_",samples[7])]],obj.ls_RNA[[paste0(mod,"_",samples[8])]]))

}


```

QC and bad cells filtering

Let's first plot some QCs 

```{r}
# The [[ operator can add columns to Oligoect metadata. This is a great place to stash QC stats
combined.obj_RNA[["RNA_AAAAGGGG"]][["percent.mt"]] <- PercentageFeatureSet(combined.obj_RNA[["RNA_AAAAGGGG"]], pattern = "^mt-")
```

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the Oligoect, i.e. columns in Oligoect metadata, PC scores etc.

plot1 <- FeatureScatter(combined.obj_RNA[["RNA_AAAAGGGG"]], feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(combined.obj_RNA[["RNA_AAAAGGGG"]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

```



Final subsetting

```{r}
combined.obj_RNA[["RNA_AAAAGGGG"]] <- subset(combined.obj_RNA[["RNA_AAAAGGGG"]], subset = nFeature_RNA > 200 & nFeature_RNA < 9000 & percent.mt < 5)
```

QCs value after subsetting 

```{r}
# Visualize QC metrics as a violin plot

QC_RNA <- data.frame(sample =  combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data[["orig.ident"]],
                     Genes = combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data[["nFeature_RNA"]],
                     RNA = combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data[["nCount_RNA"]],
                     Mt = combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data[["percent.mt"]])
QC_RNA$sample <- as.factor(QC_RNA$sample)

# Function to add mean and median annotations for each sample above the violin plots
add_mean_median_annotation <- function(plot, data, x_var, y_var) {
  samples <- unique(data[[x_var]])
  for (sample in samples) {
    sample_data <- data[data[[x_var]] == sample, ]
    mean_value <- mean(sample_data[[y_var]], na.rm = TRUE)
    median_value <- median(sample_data[[y_var]], na.rm = TRUE)
    plot <- plot + 
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.05, 
               label = paste("Mean:", round(mean_value, 2)), color = "blue", size = 5, fontface = "bold") +
      annotate("text", x = sample, y = max(sample_data[[y_var]], na.rm = TRUE) * 1.50, 
               label = paste("Median:", round(median_value, 2)), color = "red", size = 5, fontface = "bold")
  }
  return(plot)
}

# Plot for RNA
Transcripts <- ggplot(QC_RNA) +
  aes(x = sample, y = RNA, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+#,, , "#00FFBB","#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+  #, , )) +
  scale_color_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+# "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+ #, , "#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088")) +
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of UMI per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Transcripts <- add_mean_median_annotation(Transcripts, QC_RNA, "sample", "RNA")

# Plot for Genes
Genes <- ggplot(QC_RNA) +
  aes(x = sample, y = Genes, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+#,, , "#00FFBB","#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+  #, , )) +
  scale_color_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+# "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+ #, , "#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088")) +
  ggthemes::theme_base() +
  xlab("") +
  ylab("Number of Genes per cell") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Genes <- add_mean_median_annotation(Genes, QC_RNA, "sample", "Genes")

# Plot for Mt
Mt <- ggplot(QC_RNA) +
  aes(x = sample, y =   Mt, fill = sample, color = sample) +
  geom_violin(trim = TRUE, color = "#000000") +
  geom_boxplot(width = 0.1, color = "#000000", fill = "#ffffff", outlier.shape = NA) +
  scale_fill_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+#,, , "#00FFBB","#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+  #, , )) +
  scale_color_manual(values = c(
"#E41A1C",  # Red
  "#377EB8",  # Blue
  "#4DAF4A",  # Green
  "#984EA3",  # Purple
  "#FF7F00",  # Orange
  "#FFFF33",  # Yellow
  "#A65628",  # Brown
  "#F781BF"   # Pink
))+# "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088"))+ #, , "#00D3FF", "#005DFF", "#1900FF", "#8A00FF", "#FA00F9", "#FF0088")) +
  ggthemes::theme_base() +
  xlab("") +
  ylab("Percentage of Mt genes") +
  theme(panel.border = element_rect()) +
  theme_classic() + theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 18, family = "Arial", angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, family = "Arial"),
        axis.title.y = element_text(size = 18, family = "Arial"),
        axis.line = element_line(size = 1))

Mt <- add_mean_median_annotation(Mt, QC_RNA, "sample", "Mt")

Transcripts
Genes
Mt

plots <- list(UMI_Embryos=Transcripts,Genes_Embryos=Genes,Mt_Embryos=Mt)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/proj/user/mattia/Embryo_2/plots/", paste0(nm, ".png")), plots[[nm]], width = 14, height = 16, dpi = 300)
}
#how many cells 

cells_numer<-as.data.frame(table(combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data[["orig.ident"]]))

```

2. Dimensionality reduction and clustering 


First we will perform the analysis without data integration 

```{r}
combined.obj_RNA[["RNA_AAAAGGGG"]] <- NormalizeData(combined.obj_RNA[["RNA_AAAAGGGG"]])
combined.obj_RNA[["RNA_AAAAGGGG"]] <- FindVariableFeatures(combined.obj_RNA[["RNA_AAAAGGGG"]])
combined.obj_RNA[["RNA_AAAAGGGG"]] <- ScaleData(combined.obj_RNA[["RNA_AAAAGGGG"]])
combined.obj_RNA[["RNA_AAAAGGGG"]] <- RunPCA(combined.obj_RNA[["RNA_AAAAGGGG"]])
```
elbow plot

```{r}
p1<-ElbowPlot(combined.obj_RNA[["RNA_AAAAGGGG"]])
p1
```

Clustering without data integration


```{r}
combined.obj_RNA[["RNA_AAAAGGGG"]] <- FindNeighbors(combined.obj_RNA[["RNA_AAAAGGGG"]], dims = 1:15, reduction = "pca")
combined.obj_RNA[["RNA_AAAAGGGG"]] <- FindClusters(combined.obj_RNA[["RNA_AAAAGGGG"]], resolution = 2, cluster.name = "unintegrated_clusters")
```

```{r}
combined.obj_RNA[["RNA_AAAAGGGG"]] <- RunUMAP(combined.obj_RNA[["RNA_AAAAGGGG"]], dims = 1:15, reduction = "pca", reduction.name = "umap.unintegrated")
# visualize by batch and cell type annotation
# cell type annotations were previously added by Azimuth
colors <- c(
  "#E41A1C",  # red
  "#377EB8",  # blue
  "#4DAF4A",  # green
  "#984EA3",  # purple
  "#FF7F00",  # orange
  "#FFFF33",  # yellow
  "#A65628",  # brown
  "#F781BF"   # pink
)
p1<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]], reduction = "umap.unintegrated", group.by = c("unintegrated_clusters"))
p2<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]], reduction = "umap.unintegrated", group.by = c("orig.ident"),cols = colors)
p1
p2
```


Now clustering with integration based on SCTtransform will be performed 

```{r}
options(future.globals.maxSize = 140e+09)
combined.obj_RNA[["RNA_AAAAGGGG"]] <- SCTransform(combined.obj_RNA[["RNA_AAAAGGGG"]])
combined.obj_RNA[["RNA_AAAAGGGG"]] <- RunPCA(combined.obj_RNA[["RNA_AAAAGGGG"]], npcs = 30, verbose = F)
combined.obj_RNA[["RNA_AAAAGGGG"]] <- IntegrateLayers(
  object = combined.obj_RNA[["RNA_AAAAGGGG"]],
  method = RPCAIntegration,
  normalization.method = "SCT",
  verbose = F
)
combined.obj_RNA[["RNA_AAAAGGGG"]] <- FindNeighbors(combined.obj_RNA[["RNA_AAAAGGGG"]], dims = 1:15, reduction = "integrated.dr")
combined.obj_RNA[["RNA_AAAAGGGG"]] <- FindClusters(combined.obj_RNA[["RNA_AAAAGGGG"]], resolution = 0.9,cluster.name = "rpca_SCT_clusters")
```

Perform dimensionality reduction and plot it 


```{r}
combined.obj_RNA[["RNA_AAAAGGGG"]]  <- RunUMAP(combined.obj_RNA[["RNA_AAAAGGGG"]] , dims = 1:15, reduction = "integrated.dr",reduction.name = "umap.rpca_SCT")
```
plot UMAP


```{r}
colors <- c(
  "#E41A1C", # Red
  "#377EB8", # Blue
  "#4DAF4A", # Green
  "#984EA3", # Purple
  "#FF7F00", # Orange
  "#FFFF33", # Yellow
  "#A65628", # Brown
  "#F781BF"  # Pink
)

p1<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]], reduction = "umap.rpca_SCT", group.by = c("rpca_SCT_clusters"))
p2<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]], reduction = "umap.rpca_SCT", group.by = c("orig.ident"),cols = colors)


plots <- list(UMAP_Clusters=p1,UMAP_samples=p2)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/proj/user/mattia/Embryo_2/plots/", paste0(nm, ".png")), plots[[nm]], width = 20, height = 16, dpi = 300)
}
p1
p2
```

##Labeling clusters based on RNA using developing mouse brain ATLAS from La Manno et al., 2021



```{r}
# load developing embryonic brain reference atlas
embryo_brain_atlas<- readRDS("/date/gcb/gcb_MZ/multiNanoCT/Seurat_MouseDev_LaManno2021.rds")

# Assuming 'your_seurat_object' is your original Seurat object
# and 'desired_number_of_cells' is the number of cells you want to subsample
#desired_number_of_cells <- 100000
# Subsample the Seurat object
embryo_brain_atlas_subsampled <- subset(x = embryo_brain_atlas, cells = WhichCells(embryo_brain_atlas, expression = Age==c("e11.0","e12.0","e12.5","e13.0","e13.5", "e14.0","e14.5","e15.0","e15.5","e16.0")))

```




```{r}
transfer_anchors <- FindTransferAnchors(reference =embryo_brain_atlas_subsampled, query = combined.obj_RNA[["RNA_AAAAGGGG"]],
                                            reference.assay = "RNA",
                                        reduction = "cca",
                                        npcs = 50,
                                        dims = 1:50,
  query.assay = "RNA",
  features = VariableFeatures(combined.obj_RNA[["RNA_AAAAGGGG"]]@assays[["RNA"]]),)
  

combined.obj_RNA[["RNA_AAAAGGGG"]]<- MapQuery(
  anchorset =transfer_anchors, 
  reference = embryo_brain_atlas_subsampled,
  query =  combined.obj_RNA[["RNA_AAAAGGGG"]],
  refdata = "Class")

cn <- colnames(combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_cell_cluster"
cn[cn == "predicted.id.score"] <- "predicted.id.score_cluster"
colnames(combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data) <- cn

combined.obj_RNA[["RNA_AAAAGGGG"]]<- MapQuery(
  anchorset =transfer_anchors, 
  reference = embryo_brain_atlas_subsampled,
  query =  combined.obj_RNA[["RNA_AAAAGGGG"]],
  refdata = "Age")

cn <- colnames(combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data)
cn[cn == "predicted.id"] <- "predicted.id_Age"
cn[cn == "predicted.id.score"] <- "predicted.id.score_Age"
colnames(combined.obj_RNA[["RNA_AAAAGGGG"]]@meta.data) <- cn


```

```{r}
colors <- c(
  "#E41A1C", # Red
  "#377EB8", # Blue
  "#4DAF4A", # Green
  "#984EA3", # Purple
  "#FF7F00", # Orange
  "#FFFF33", # Yellow
  "#A65628", # Brown
  "blue", # Pink
  "#999999", # Grey
  "#66C2A5", # Teal
  "black", # Salmon
  "#8DA0CB", # Light Blue
  "#E78AC3", # Light Pink
  "#A6D854", # Lime
  "#FFD92F" # Bright Yellow
)

p1<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",group.by = "predicted.id_cell_cluster",cols = colors)
p2<-FeaturePlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",features = "predicted.id.score_cluster")
p3<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",group.by = "predicted.id_Age",cols = colors)
p4<-FeaturePlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",features = "predicted.id.score_Age")
p1
p2
p3
p4

plots <- list(UMAP_labels=p1,UMAP_labels_score=p2)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/proj/user/mattia/Embryo_2/plots/", paste0(nm, ".png")), plots[[nm]], width = 20, height = 16, dpi = 300)
}


#saveRDS(combined.obj_RNA[["RNA_AAAAGGGG"]],"/proj/user/mattia/Embryo_2/combined.obj_RNA_embryos.rds")
```

dot plot markers

```{r , fig.height=10, fig.width=20}
markers<-c("Sox2","Pax6","Nes"," Slc1a3","Neurog2","Bmp4","Bmp7","Mki67","Eomes",#Radial Glia
           "Dcx", "Satb2","Tbr1","Rbfox3","Dlx2","Dlx5", #Neu
           "Pdgfra","Sox9","Sox6","Sox10","Egfr","Gfap",#Glia
           "Mag","Plp1"
           
)

p1<-DotPlot(combined.obj_RNA[["RNA_AAAAGGGG"]], features = markers,group.by = "predicted.id_cell_cluster",cols = c("lightgrey", "firebrick"))+
  xlab("")+
  theme(
    axis.text.x = element_text( hjust = 1, vjust = 0.5, family = "Arial", size = 20,angle = 90),
    axis.text.y = element_text(family = "Arial", size = 20),
    axis.title.y = element_text(family = "Arial", size =24),
    legend.text = element_text(size = 20,family = "Arial"),
    legend.title = element_text(size = 24,family = "Arial")
  ) 
p2<-FeaturePlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",features = markers)
plots <- list(DotPlot_markers=p1)
for (nm in names(plots)) {
ggplot2::ggsave(file.path("/proj/user/mattia/Embryo_2/plots/", paste0(nm, ".png")), plots[[nm]], width =14, height = 16, dpi = 300)
}
p1
p2
```


Renaming clusters

```{r}
colors <- c(
  "#E6194B", # Red
  "#3CB44B", # Green
  "yellow", # Yellow
  "#0082C8", # Blue
  "#F58231", # Orange
  "#911EB4", # Purple
  "#46F0F0", # Cyan
  "#F032E6", # Magenta
  "#D2F53C", # Lime
  "#FABEBE", # Pink
  "#008080", # Teal
  "#E6BEFF", # Lavender
  "#AA6E28", # Brown
  "#FFFAC8", # Beige
  "#800000", # Maroon
  "#AAFFC3", # Mint
  "#808000", # Olive
  "#FFD8B1", # Apricot
  "#000080", # Navy
  "#808080", # Grey
  "purple", # White
  "#000000", # Black
  "#00FFFF", # Turquoise
  "#FFD700" , # Gold
  "green"
)

p1<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",group.by = "seurat_clusters",cols=colors)
p2<-DimPlot(combined.obj_RNA[["RNA_AAAAGGGG"]],reduction = "umap.rpca_SCT",group.by = "predicted.id_cell_cluster",cols = colors)
p1
p2
```

```{r}
# Suppose your clusters are "0", "1", "2", ...
# and you want to rename them as follows:
new.names <- c("0" = "Neurons", "1" = "Radial Glia", "2" = "Glia precursors", "3" = "Neurons",
               "4"="Neurons","5"="Intermediate Progenitors","6"="Neurons","7"="Intermediate Progenitors",
               "8"="Neurons","9"="Intermediate Progenitors","10"="Fibroblast","11"="Neurons",
               "12"="Glioblast","13"="Fibroblast","14"="Neurons 2", "15"="Schwann Cells",
               "16"="Vascular","17"="Blood","18"="Radial Glia 2","19"="Neurons",
               "20"="Fibroblast","21"="Neurons","22"="immune","23"="Glia precursors",
               "24"="Choroid Plexus")

combined.obj_RNA[["RNA_AAAAGGGG"]] <- RenameIdents(combined.obj_RNA[["RNA_AAAAGGGG"]], new.names)
combined.obj_RNA[["RNA_AAAAGGGG"]]$cluster_renamed <- Idents(combined.obj_RNA[["RNA_AAAAGGGG"]])

saveRDS(combined.obj_RNA[["RNA_AAAAGGGG"]],"/proj/user/mattia/Embryo_2/combined.obj_RNA_embryos_high_quality.rds")

```
















